The E3 ubiquitin ligase RNF220 maintains hindbrain Hox expression patterns through regulation of WDR5 stability

Version 1

Abstract

Abstract

The spatial and temporal linear expression of Hox genes establishes a regional Hox code, which is crucial for the anteroposterior patterning, segmentation, and neuronal circuit development of the hindbrain. RNF220, an E3 ubiquitin ligase, is widely involved in neural development via the targeting of multiple substrates. Here, we found that the expression of Hox genes in the pons was markedly up-regulated at the late developmental stage (post-embryonic day E15.5) in Rnf220−/− and RnfF220+/− mouse embryos. Single-nucleus RNA-seq analysis revealed different Hox de-repression profiles in different groups of neurons, including the pontine nuclei (PN). The Hox pattern was disrupted and the neural circuits were affected in the PN of Rnf220+/− mice. We showed that this phenomenon was regulated by WDR5, a key component of the TrxG complex, which can be ubiquitinated and degraded by RNF220. Intrauterine injection of WDR5 inhibitor (WDR5-IN-4) and genetic ablation of Wdr5 in Rnf220+/− mice largely recovered the de-repressed Hox expression pattern in the hindbrain. In P19 embryonal carcinoma cells, the retinoic acid induced Hox expression was also stimulated upon Rnf220 knockdown, which can be further rescued by Wdr5 knockdown. In short, our data suggest a new role of RNF220/WDR5 in Hox pattern maintenance and pons development.

Introduction

The coordinated expression of Hox genes clusters is critical for axial patterning to determine the positional identities during the development of many tissues, including the hindbrain (Frank et al., 2019; Parker et al., 2014; Parker et al., 2020; Pöpperl et al., 1993.). In mammals, there are 39 Hox genes organized in four clusters (HoxA-HoxD) and divided into 1-13 paralog groups (PG). During vertebrate development, the hindbrain forms eight metameric segmented units along the antero-posterior (A-P) axis known as rhombomeres (r), which later give rise to the cerebellum, pons and medulla later on (Ghosh et al., 2018; Parker et al., 2020.). In the embryonic hindbrain, Hox1-5 are expressed in a nested overlapping fashion with rhombomere-specific boundaries. This Hox code is critical for rhombomeric territories establishment and cell fate determination (Barsh et al., 2017; Parker et al., 2016; Parker et al., 2020.).

At later stages, the Hox genes also play vital roles in neuronal migration and neural circuit formation (Feng et al., 2021; Hockman et al., 2019; Kratochwil et al., 2017; X. Wang et al., 2013.). For example, the pontine nuclei (PN) in the pons acts as an information integration station between the cortex and cerebellum (Kratochwil et al., 2017; Maheshwari et al., 2020.). The PN neurons originate at the posterior rhombic lip of r6-r8 and migrate tangentially to the rostral ventral hindbrain (Di Meglio et al., 2008; Farago et al., 2006; Geisen et al., 2008; Kratochwil et al., 2017; Maheshwari et al., 2020.). The PN neurons express Hox 3-5 in a nested pattern determined by their rhombomeric origin, and maintained throughout migration and PN assembly (Di Meglio et al., 2013; Kratochwil et al., 2017; Lizen et al., 2017.). Hox factors are crucial for the specification, migration, cellular organization, and topographic input connectivity of PN neurons. Note that Hox gene expression is maintained up to postnatal and even adult stages in the hindbrain derivatives (Farago et al., 2006; Feng et al., 2021; Kratochwil et al., 2017.).

Hox pattern maintenance is generally governed by epigenetic regulators, especially the polycomb-group (PcG) and trithorax-group (trxG) complexes, which activate and repress Hox genes, respectively (Bahrampour et al., 2019; Chopra et al., 2009; Kang et al., 2022; Papp et al., 2006.). In mammals, the COMPASS (complex of proteins associated with Set1) family of TrxG complexes act as methyltransferases. They contain different SET/MLL proteins, and a common multi-subunit core module consisting of WDR5, RbBP5, Ash2L, and DPY-30 (WRAD) (Cenik et al., 2021; Jambhekar et al., 2019; Schuettengruber et al., 2017.). WDR5 plays a central scaffolding role in these complexes and has been shown to act as a key regulator of Hox maintenance (K. C. Wang et al., 2011; Wysocka et al., 2005.). It is involved in MLL-mediated HoxA9 activation in acute myeloid leukemia (AML) and acute lymphoblastic leukemia (ALL) and is a target for drug development (Chen et al., 2021; Yu et al., 2021.). The long noncoding RNA (lncRNA) HOTTIP also binds to WDR5 and recruits the MLL complex to stimulate HoxA9 and HoxA13 expression in prostate cancer and hepatocellular carcinoma (Fu et al., 2017; Malek et al., 2017; Quagliata et al., 2014; Wong et al., 2020.).

RNF220 is an E3 ubiquitin ligase widely involved in neural development through targeting a range of proteins (Kim et al., 2018; Kong et al., 2010; Pengcheng Ma & Mao, 2021; Pengcheng Ma et al., 2022; Y.-B. Wang et al., 2022.). Previous research has shown that Rnf220−/− mice exhibit neonatal lethality and Rnf220+/− mice develop severe motor impairments; however, the mechanisms responsible for these phenotypes remain incompletely understood (Pengcheng Ma, Li, et al., 2021; Pengcheng Ma et al., 2019.). In the present study, we observed marked up-regulation of Hox genes in the pons of both Rnf220+/−and Rnf220−/− mice during late embryonic development. In the PN of Rnf220+/− mice, both the Hox expression pattern and the projection pattern from the motor cortex to the PN were disrupted. We identified WDR5 as a direct target of RNF220 for K48-linked ubiquitination and thus degradation, a mechanism critical for Hox regulation by RNF220. These findings revealed a novel role for RNF220 in the regulation of Hox genes and pons development.

Results

Rnf220 insufficiency leads to dysregulation of Hox expression in late embryonic pons

Rnf20 is strongly expressed in the embryonic mouse mid/hind brain (Pengcheng Ma et al., 2019; Y.-B. Wang et al., 2022.). Based on microarray analysis, we explored changes in expression profiles in E18.5 Rnf220+/−and Rnf220−/− mouse brain. Interestingly, several Hox genes were up-regulated in the mouse brain of both genotypes (Tables 1 and 2). Considering that Hox expression in forebeain is to low (date not shown), we focus on hindbrain and found that this phenomenon was observed in late embryonic stages after E15.5 but not earlier (Figure 1-figure supplement 1). To determine the specific brain regions exhibiting aberrant Hox expression, we examined the expression levels of HoxA9 and HoxB9, which were among the most significantly up-regulated genes, across various brain and spinal regions (Figure 1-figure supplement 2A-D). Notably, both HoxA9 and HoxB9 were exclusively up-regulated in the brainstem (Figure 1-figure supplement 2C and D). The mouse brainstem comprises the midbrain, pons, and medulla (Figure 1-figure supplement 2E). Further validation confirmed that the dysregulation of these Hox genes was restricted to the pons in the Rnf220+/− mice (Figure 1-figure supplement 2F). RNA sequencing (RNA-seq) analysis of the pons in 2-month-old Rnf220+/−mice similarly revealed an overall increase in Hox expression (Figure 1A).

Up-regulation of Hox genes in pons of Rnf220+/− mice. (A) The heatmap of RNA-seq data showing Hox genes expression in pons of WT or Rnf220+/− mice (n=2 mice per group). (B-C) UMAP diagram showing 15 identified cell clusters annotated by snRNA-seq analysis of pons. Each dot represents a single cell, and cells are laid out to show similarities (n=3 mice per group). (D) Heatmap of snRNA-seq data showing Hox expression changes in each cell cluster. (E) Quantitative real time polymerase chain reaction, qRT-PCR) analysis showing mRNA levels of indicated Hox genes in P19 cells when endogenous Rnf220 was knocked down by siRNAs in the presence of RA. WT, wild-type. HE, heterozygote. RA, retinoic acid. **p < 0.01, ***p < 0.001.

To identify the specific cell population in which Hox genes were up-regulated, we conducted single-nucleus RNA-seq (snRNA-seq) analysis of the pons in 2-month-old wild-type (WT) and Rnf220+/− mice. In total, 125 956 cells were resolved in the WT mice and 122 266 cells were resolved in the Rnf220+/− mice, each containing an average of approximately 700 genes per cell. Subsequent uniform manifold approximation and projection (UMAP) identified 15 cell clusters, with no significant difference observed between the WT and Rnf220+/− groups (Figure 1B and C). These clusters were annotated based on their uniquely and highly expressed markers (Table 3). Among them, 10 clusters were identified as distinct neuronal groups, in addition to astrocytes, oligodendrocytes, oligodendrocyte precursors, ependymal cells, and a group not corresponding to any known cell type (Figure 1B). We analyzed changes in Hox gene expression within each cluster and found that up-regulation of Hox genes was most pronounced in clusters 2, 7, and 10 (Figure 1D), albeit with distinct profiles. For instance, Hox8–10 were the most up-regulated genes in cluster 2, while Hox3–5 and Hox1–3 were the most up-regulated genes in clusters 7 and 10, respectively (Figure 1D). These findings indicate that Hox genes were specifically de-regulated in the pons of Rnf220+/− mice during late developmental stages.

P19 embryonal carcinoma cells can be induced to differentiate and express Hox genes upon retinoic acid (RA) administration and have been used to study Hox regulation (Figure 1-figure supplement 3A)(Heyden et al., 2003; Kondo et al., 1992; Pöpperl et al., 1993.). Here, we tested the effects of Rnf220 knockdown on RA-induced Hox expression in P19 cells. Without RA treatment, Rnf220 knockdown had no induction effect on Hox expression (Figure 1-figure supplement 3B). However, with RA treatment, Rnf220 knockdown increased RA-induced expression of HoxA1, B1, A9, and B9 (Figure 1E), suggesting a general role of Rnf220 in Hox regulation.

Rnf220 insufficiency disturbs Hox expression pattern and circuit formation in PN

Due to limitations in data resolution, the precise identification of each cluster was not fully resolved. We determined that cluster 10 likely represents cells derived from rhombomeres 2–5, given their expression of endogenous anterior Hox genes (Hox1–3), while cluster 7 likely represents cells derived from rhombomeres 6–8 cells, as evidenced by their expression of Hox3–5 (Figure 1D). After mapping known neuron-specific markers to the clusters, PN-specific markers, including Nfib, Pax6, and Barhl, were enriched in cluster 7 (Figure 2-figure supplement 1A and B). PN neurons serve as relay cells between the cerebral cortex and cerebellum, originating at the posterior rhombic lip and migrating tangentially to their final position in the ventral part of the pons (Di Meglio et al., 2013; Kratochwil et al., 2017; Maheshwari et al., 2020.). Hox genes are implicated in both PN patterning and circuit formation (Kratochwil et al., 2017; Maheshwari et al., 2020.). In this study, we explored the potential effects of Rnf220 insufficiency on PN patterning. The PN were first dissected to analyze the expression of Hox genes (Hox4 and Hox5) in both WT and Rnf220+/− mice (Figure 2-figure supplement 1C). The data indicated an up-regulation in Hox4–5 expression levels in the Rnf220+/− PN (Figure 2-figure supplement 1C). In the PN, endogenous Hox3–5 genes exhibit a nested expression pattern along the rostral-caudal axis, with Hox3 ubiquitously expressed throughout the PN, Hox4 localized to the middle and posterior regions, and Hox5 confined to the posterior segment (Kratochwil et al., 2017.). To ascertain whether this established pattern was disrupted in the Rnf220+/− PN, we dissected the PN of both WT and Rnf220+/− mice into rostral, middle, and caudal segments using a vibratome. Subsequent analysis of Hox3–5 expression revealed an up-regulation in Rnf220+/−PN, with uniform detection along the rostral-caudal axis (Figure 2A).

Hox gene expression was dysregulated and motor cortex projections were disorganized in PN of Rnf220+/− mice. (A) qRT-PCR analysis of relative expression levels of Hox3, Hox4, and Hox5 in rostral, middle, and caudal sections of PN in WT and RNF220+/− mice. Expression level of each gene in rostral section of WT PN was set to 1. (n=4 mice per group). (B) Diagram of experimental stereotactic injections. (C, D) Fluorescent signals showing projection from motor cortex to PNs in WT and Rnf220+/− mice (n=2 mice per group). WT, wild-type; HE, heterozygote; **p < 0.01; ***p< 0.001.

PN are predominantly innervated by the cerebral cortex and Hox expression is critical for the formation of proper projection patterns (Kratochwil et al., 2017; Maheshwari et al., 2020.). Considering the motor deficits of Rnf220+/− mice (Pengcheng Ma, Li, et al., 2021; Pengcheng Ma et al., 2019.), we next tested whether projection patterns were affected by anterograde tracing. A non-transneuronal tracing virus, rAAV-hSyn-EGFP-WPRE-hGH-poly A, was injected into the motor cortex of 2-month-old WT and Rnf220+/−mice to label axons sent from the motor cortex (Figure 2B). Green fluorescence intensity in the PN was measured after 3 weeks. In the PN, the projection range from the motor cortex was smaller in Rnf220+/− mice (Figure 2C and D). In addition, the projection pattern was shifted so that more fluorescence intensity was observed in the middle section in the Rnf220+/− PN compared to the wild type (Figure 2C and D).

Thus, Rnf220 insufficiency disturbed both the expression level and pattern of Hox genes in PN and affected the projection pattern from the motor cortex.

WDR5 is an RNF220 target for K48-linked ubiquitination and degradation

As Hox up-regulation was observed only in the late stage of embryonic development, we reasoned that Rnf220 may be involved in the maintenance of Hox expression. The Hox expression pattern is maintained epigenetically by the PcG and trxG complexes, which regulate the silencing and activation of Hox genes, respectively (Bahrampour et al., 2019; Chopra et al., 2009; Kang et al., 2022; Papp et al., 2006.). Indeed, we did observe local epigenetic modification changes in each Hox cluster in hindbrain of Rnf220+/− mice by ChIP-qPCR assays (Figure 3-figure supplement 1A and B). Therefore, we tested the expression of core components of PRC1/2 in the hindbrain (E18.5) and pons (2 months old) of Rnf220+/−mice and found no obvious change in protein levels (Figure 3-figure supplement 2A and B). However, when WDR5, a key component of the trxG complex, was tested, a clear increase was observed in the hindbrain of both Rnf220+/−and Rnf220−/− mouse embryos at E18.5 (Figure 3A), which was already evident at E16.5 (Figure 3B). We next tested the expression of WDR5 in the pons, cortex, and cerebellum of 2-month-old WT and Rnf220+/− mice. WDR5 was up-regulated in the pons but not in the cortex or cerebellum (Figure 3C, Figure 3-figure supplement 2C and D), consistent with the pons-specific Hox up-regulation. In addition, WDR5 also increased in the PN of Rnf220+/− mice (Figure 3D) and RA-treated P19 cells following Rnf220 knockdown (Figure 3E). Thus, the stabilization of WDR5 was correlated with Rnf220-regulated Hox expression.

RNF220 mediates WDR5 degradation. (A-D) Western blots analysis showing the protein level of WDR5 in the indicated brain tissues of mice with different genotypes at different ages. (E) Western blot analysis of protein levels of WDR5 in P19 cells with Rnf220 knockdown or not in the presence or absence of RA. IB, immunoblot. WT, wild-type. HE, heterozygote. KO, knockout. PN, pontine nuclei. NC, negative control. RA, retinoic acid.

We speculated that WDR5 might be a direct target of RNF220 for ubiquitination and degradation. Indeed, WDR5 efficiently pulled down endogenous RNF220 in mouse brain stem lysate (Figure 4A), and RNF220 and WDR5 pulled down each other when co-expressed in HEK293 cells (Figure 4B and C). Furthermore, co-expression of wild-type RNF220 but not the ligase-dead mutant (W539A) or the RING domain deletion (△Ring) form clearly reduced the expression level of WDR5 in HEK293 cells (Figure 4D), which was blocked by MG132, a proteasome inhibitor (Figure 4E). In ubiquitination assays, WDR5 ubiquitination was reduced in the lysate of the pons of Rnf220+/−mice (Figure 4F). Accordingly, the level of ubiquitinated WDR5 clearly increased when co-expressed with RNF220 but not the △Ring form in HEK293 cells (Figure 4G).

RNF220 interacts with and targets WDR5 for K48-linked polyubiquitination. (A) Endogenous co-immunoprecipitation analysis showing the interaction between RNF220 and WDR5 in hindbrains of WT mice. (B-C) Co-immunoprecipitation (co-IP) analysis of interactions between RNF220 and WDR5 in HEK293 cells. HEK293 cells were transfected with indicated plasmids and harvested after 48 h. Cell lysates were immunoprecipitated with anti-FLAG beads. Whole-cell lysate and immunoprecipitates were subjected to western blot analysis using indicated antibodies. (D) Western blots analysis shows the protein level of WDR5 when co-expressed with wild type or mutated RNF220 in HEK293 cells. (E) Western blots analysis shows the protein level of WDR5 when co-expressed with RNF220 in HEK293 cells in the presence of MG132 or not. MG132 treatment reversed the RNF220-induced decrease of WDR5 protein. (F) In vivo ubiquitination assays showing the ubiquitination status of WDR5 in hindbrains of WT or Rnf220+/− mice. (G) In vivo ubiquitination assays showing the ubiquitination status of WDR5 when co-expressed with WT or mutated RNF220 in HEK293 cells. (H) In vivo ubiquitination assays showing RNF220-induced polyubiquitination of WDR5 when the indicated ubiquitin mutations were used in HEK293 cells. (I) In vivo ubiquitination assays showing the ubiquitination status of the indicated WDR5 mutants when co-expressed with WT or ligase dead RNF220 in HEK293 cells. WT, wild-type. HE, heterozygote. KO, knockout. IB, immunoblot. UB, ubiquitin. WCL, whole cell lysate. △Ring, RNF220 Ring domain deletion. W539R, RNF220 ligase dead mutation. K48, ubiquitin with all lysines except the K48 mutated to arginine. K48R, ubiquitin with the K48 was substituted by an arginine. 3M, substitution of lysines at the positions of 109, 112, and 120 in WDR5 with arginines simultaneously.

We tested whether the polyubiquitination of WDR5 by RNF220 was K48 linked, which is the major form of ubiquitination modification leading to proteasomal degradation (Grice et al., 2016.). Indeed, the K48 ubiquitin mutant (in which all the lysines are substituted by arginines except lysine 48) was efficiently incorporated into the ubiquitinated WDR5 induced by RNF220, but not the K48R mutant (in which lysine 48 was substituted by an arginine), confirming that the polyubiquitination chains are K48-linked (Figure 4H). To determine the exact lysines ubiquitinated by RNF220, we tested the effects of RNF220 on the stability of different WDR5 truncates and found that the truncate remaining 1-127aa was enough to be degraded by RNF220 (Figure 4-figure supplement 1A). We then mutated the conserved lysines in this region to arginine individually, and found that mutation of K109, K112, or K120 reduced the ubiquitination level of WDR5 by RNF220 (Figure 4-figure supplement 1B). Simultaneous substitution of these three lysines with arginine (3M) greatly reduced WDR5 ubiquitination by RNF220 (Figure 4I). In conclusion, these findings indicate that RNF220 mediates K48-linked polyubiquitination of WDR5 primarily at residues K109, K112, and K120.

WDR5 is required for Hox regulation by RNF220

To investigate the requirement of WDR5 in RNF220-mediated Hox regulation, we first examined whether WDR5 knockdown could mitigate the impact of Rnf220 knockdown on Hox expression in RA-induced P19 cells. When P19 cells were transfected with siWdr5 together with siRnf220 and accompanied by RA treatment, the stimulation of Hox genes by RNF220 knockdown was significantly reduced (Figure 5A-D). Note that the expression of Hox genes had no clear change upon Wdr5 knockdown alone with or without RA treatment in P19 cells (Figure 5-figure supplement 1).

WDR5 recovered Rnf220 deficiency-induced up-regulation of Hox genes in P19 cell line. (A-B) qRT-PCR analysis showing the expression levels of Rnf220 (A) and Wdr5 (B) when transfected the indicated combinations of siRNAs against Rnf220 or Wdr5 in the presence or absence of RA. (C-D) qRT-PCR analysis showing the expression levels of HoxA1, HoxB1, HoxA9, HoxB9 when transfected siRnf220 or both siRnf220 and siWdr5 when treated with RA (D) or not (C). RA: retinoic acid. n.s., not significant. *p< 0.05, **p < 0.01, ***p < 0.001.

We also tested in mice. we first examined the potential for WDR5-IN-4, an established WDR5 inhibitor (Aho et al., 2019.), to rescue Hox up-regulation in Rnf220+/− embryos. Using in-utero microinjection, we injected WDR5-IN-4 or DMSO in mice at E15.5 and harvested the embryos at E18.5, with the hindbrains processed for qRT-PCR analysis (Figure 6A). Results showed that WDR5-IN-4 had no effect on RNF220 expression (Figure 6B). Interestingly, the up-regulation of HoxA9 and HoxB9 in Rnf220+/− mice was reversed in the WDR5-IN-4 injected embryos (Figure 6B).

Genetic and pharmacological ablation of WDR5 recovered Rnf220 deficiency-induced up-regulation of Hox genes. (A) Diagram of experimental strategy for in utero local injection of WDR5 inhibitors. (B) qRT-PCR analysis of expression levels of Rnf220, HoxA9, and HoxB9 in hindbrains of WT and Rnf220+/− mouse embryos treated with WDR5 inhibitors or not at E18.5 (n=3 mice per group). (C) qRT-PCR analysis of expression levels of Rnf220, Wdr5, HoxA9, and HoxB9 in hindbrains of adult mice with indicated genotypes (n=2 mice per group). WT, wild-type. HE, heterozygote. ***p < 0.001.

As loss of RNF220 stabilizes WDR5, we next tested whether WDR5 insufficiency would rescue the Hox expression pattern in Rnf220+/−;Wdr5+/−double heterozygous embryos. Rnf20fl/wt;Nestin-Cre mice were mated with Wdr5fl/wt mice to obtain embryos with different genotypes. The expression of Hox genes in the pons at P15 were analyzed by RT-PCR (Figure 6C, Figure 6-figure supplement 1). Results showed that the up-regulation of Hox in Rnf20fl/wt;Nestin-Cre mice were markedly recovered in the Rnf220fl/wt;Wdr5fl/wt;Nestin-Cre mice (Figure 6C, Figure 6-figure supplement 1).

Taken together, the above findings support the involvement of WDR5 in RNF220 insufficiency-induced Hox up-regulation.

Discussion

During early embryonic development, the Hox genes are collinearly expressed in the hindbrain and spinal cord along the A-P axis to guide regional neuronal identity. Later on, the segmental Hox gene expression pattern in the hindbrain is maintained till at least early postnatal stages. The Hox genes are also expressed in adult hindbrains with restricted anterior boundaries. In addition to their early roles in progenitor cell specification, cell survival, neuronal migration, axon guidance, and dendrite morphogenesis (Smith et al., 2024.), Hox genes also play key roles in the regulation of synapse formation, neuronal terminal identity and neural circuit assembly at late stages (Feng et al., 2020; Feng et al., 2021; Philippidou et al., 2013.). However, the mechanisms and functions of Hox expression during late hindbrain development remain poorly established. Our work revealed a dose-dependent role of RNF220/WDR5 in the maintenance of Hox expression in late hindbrain development which might have a functional role in the neural circuit organization of the pons.

In mammals, the COMPASS family of histone-lysine methyltransferase complexes is required for the maintenance of active chromatin states and thus Hox expression patterns (Bahrampour et al., 2019; Chopra et al., 2009; Kang et al., 2022; Papp et al., 2006.). As a key component of the COMPASS-related complexes, WDR5 is involved in the developmental regulation of Hox genes (Wysocka et al., 2005.). WDR5 is required for MLL-fusion protein (MLL-FP)-induced Hox activation in AML and ALL (Chen et al., 2021; Yu et al., 2021.). The lncRNA HOTTIP has been shown to interact with WDR5 and recruit WDR5-MLL complexes to HoxA to induce H3K4 trimethylation and gene activation (Fu et al., 2017; Malek et al., 2017; Quagliata et al., 2014; Wong et al., 2020.). Our findings suggest that increased WDR5 levels in the hindbrain were associated with aberrant Hox expression at the late developmental stages. This dysregulation could be ameliorated either by WDR5 inhibition or haploinsufficiency. The observed up-regulation of Hox genes occurred at the epigenetic level within the Hox cluster, as evidenced by qRT-PCR analysis and a generalized increase in active H3K4 methylation markers. Furthermore, the stabilization of WDR5 and stimulation of Hox expression upon RNF220 knockdown were also recapitulated in RA-treated P19 cells.

Interestingly, single-cell analysis revealed distinct profiles of Hox gene activation across different neuronal groups. Specifically, Hox1–3 was up-regulated in cluster 10, Hox3–5 was up-regulated in cluster 7, and Hox7–10 was up-regulated in cluster 2. Clusters 10 and 7 naturally expressed Hox1–3 and Hox3–5, respectively, suggesting that the observed Hox gene stimulation in these clusters could be attributed to elevated WDR5 levels induced by RNF220 insufficiency. Cluster 2 warrants particular attention, as it exhibited up-regulation of Hox7–10 genes, which were not expressed in the WT hindbrain. This cluster was characterized as glycinergic neurons expressing Slc6a5. The functional implications of this distinct cell group and the phenotypic consequences of Hox gene dysregulation require further investigation. The dysregulation of Hox genes could have extensive implications for the development of various nuclei within the pons, affecting aspects such as cell differentiation and the formation of neural circuits. For the PN, we showed that both their structural patterning and patterns of projections originating from the motor cortex were altered in Rnf220+/−mice. Therefore, we can fully suspect that the motor impairments of Rnf220+/−mice is caused by abnormal projection of the central nervous system, and of course, more convincing experimental verification are needed.

In summary, our data support WDR5 as an RNF220 target involved in the maintenance of Hox expression and, consequently, the development of the pons.

Materials and Methods

Mouse strains and genotyping

All procedures involving mice were conducted in accordance with the guidelines of the Animal Care and Use Committee of the Kunming Institute of Zoology, Chinese Academy of Sciences (IACUC-PA-2021-07-018). Mice were housed under standard conditions at a temperature range of 20–22 °C, humidity of 30%–45%, and 12-h light/dark cycle. All mice used were maintained on a C57BL/6 background. Vasa-Cre mice were used to mate with Rnf220 floxed mice (Rnf220fl/fl) to generate Rnf220 germ cell knockout (Rnf220−/−) or heterozygote (Rnf220+/−) mice. Nestin-Cre mice were used to mate with Rnf220fl/fland Wdr5 floxed mice (Wdr5fl/fl) to generate conditional neural specific knockout or heterozygote mice. Rnf220fl/wt;Wdr5fl/wt;Nestin-Cre mice were obtained by crossing Rnf220fl/wt;Nestin-Cre with Wdr5fl/wt mice.

Genotypes were determined by PCR using genome DNA from tail tips as templates. PCR primers were listed as follows: 5’-CTG TTG ATG AAG GTC CTG GTT-3’ and 5’-CAG GAA AAT CAA TAG ACA ACT T-3’ were used to detect Rnf220 floxP carrying. 5’-CTG TTG ATG AAG GTC CTG GTT-3’ and 5’-CTG ATT TCC AGC AAC CTA AA-3’ were used to detect Rnf220 knockout. 5’-GCC TGC ATT ACC GGT CGA TGC-3’ and 5’-CAG GGT GTT ATA AGC AAT CCC-3’ were used for Cre positive detection. 5’-GAA TAA CTA CTT TCC CTC AGA CC-3’ and 5’-CAG GCC AAG TAA CAG GAG GTA G-3’ were used to detect Wdr5 floxP carrying. 5’-GAA TAA CTA CTT TCC CTC AGA CC-3’ and 5’-AGA CCC TGA GTG AGG ATA CAT AA-3’ were used to detect Wdr5 knockout.

Cell culture

HEK293 and P19 cell lines were grown in Dulbecco’s Modified Eagle Medium (DMEM) (Gibco, C11995500BT) supplementary with 10% fetal bovine serum (FBS) (Biological Industries, 04-001-1A), 100 units/mL penicillin and 100 mg/ml streptomycin (Biological Industries, 03-031-1B). Cell cultures were maintained at 37°C in a humidified incubator with 5% CO2. 0.5μM RA (Sigma, R2625) was used to induce Hox expression in P19 cells.

To achieve gene overexpression or knockdown, HEK293 and P19 cells were transfected by Lipo2000 (Invitrogen, 11668500) according to the manufacturer’s instructions. The following small interfering RNAs (siRNAs) (RiboBio) were used for Rnf220 or Wdr5 knockdown in P19 cells: siG2010160325456075, siG2010160325457167, or siG2010160325458259; and siWdr5: siB09924171210, siG131113135429, or siG131113135419.

Total RNA isolation and qRT–PCR

Tissue and cells were homogenized with 1 mL TRIZOL (TIANGEN, DP424), after which 200 μL of chloroform was added to the lysates for phase separation. After centrifugation at 12 000 g for 15 min at 4 °C, the aqueous phase (500 μL) was transferred to a new tube and mixed with equal volumes of isopropanol for RNA precipitation. After 30 min, RNA pellets were harvested by centrifugation at 12 000 g for 15 min at 4 °C, twice washed with 75% ethanol, and dissolved in DNase/RNase-free water.

cDNA was synthesized with Strand cDNA Synthesis Kit (ThermoScientific, K1632) according to the manufacturer’s instructions. All reactions were performed at least triplicates with Light Cycler 480 SYBR Green Ⅰ Master (Roche, 04707516001). Primers used for real-time polymerase chain reaction list in Table 4.

ChIP-qPCR

Fresh mouse PN tissues were fully grinded, cross-linked with 1% formaldehyde for 25 min, and stopped with addition of glycine. After gentle centrifuging, samples were collected and lysed with ChIP lysis buffer (150 mM NaCl, 25 mM Tris, pH 7.5, 1% Triton X-100, 0.1% SDS, 0.5% deoxycholate, and complete protease inhibitor cocktail) for 30 min on ice. Genomic DNA was sheared into 200-500 bp fragments by sonication. After centrifugation, the supernatants containing chromatin fragments were collected and immunoprecipitated with anti-IgG (2 μg, ProteinTech, B900610), anti-H3K27me3 (2 μg, Abcam, ab192985), or anti-H3K4me3 antibodies (2 μg, Abcam, ab8580) at 4℃ overnight. Then the immunoprecipitates coupled with protein A/G agarose beads (Santa Cruz, sc-2003) were washed sequentially by a low salt washing buffer, high salt washing buffer, LiCl washing buffer, and TE buffer. The immunoprecipitated chromatin fragments were eluted by 500 μL elution buffer for reversal of cross-linking at 65 °C overnight. Input or immunoprecipitated genomic DNA was purified by the QIAquick PCR Purification Kit (Qiagen, 28104) and used as a template for quantification PCR. The primers we used were listed as follows:

In utero microinjection

Pregnant mice were administered isoflurane for deep anesthesia. Following this, a laparotomy was performed to carefully extract the embryos, which were then placed on a sterile surgical drape. WDR5-IN-4 (100 μg, MedChemExpress, HY-111753A), containing 0.05% Malachite Green reagent for tracing, was injected into the hindbrain of E15.5 embryos using a finely-tapered borosilicate needle. To minimize the risk of spontaneous abortion, injections were spaced for the selected embryos. Following the injections, the uterus was returned to the abdominal cavity and infused with 2 mL of 37 °C, 0.9% saline solution. The peritoneum and abdominal skin were then sutured. Finally, the mice were placed on a heating pad to facilitate recovery from anesthesia.

Ubiquitination assay, immunoprecipitation and western blot analysis

In vivo ubiquitination, immunoprecipitation and Western blot assays were carried out as previously described (Pengcheng Ma et al., 2023.). The following primary antibodies were used for botting: anti-RNF220 (Sigma-Aldrich, HPA027578), anti-WDR5 (D9E11) (Cell Signaling Technology, 13105), anti-Ring1B (D22F2) (Cell Signaling Technology, 5694S), anti-Sin3B (AK-12) (Santa Cruz, sc-768), anti-EZH2 (ProteinTech, 21800-1-AP), anti-SUZ12 (Bethyl, A302-407A), anti-RYBP (A-1) (Santa Cruz, sc-374235), anti-CBX6 (H-1) (Santa Cruz, sc-393040), anti-CBX7 (G-3) (Santa Cruz; sc-376274), anti-CBX8 (C-3) (Santa Cruz, sc-374332), anti-PHC1 (D-10) (Santa Cruz, sc-390880), anti-α-Tubulin (ProteinTech, 66031-1-Ig), anti-FLAG (Sigma-Aldrich; F-7425), anti-myc (ProteinTech; 16286-1-AP), and anti-Ub (P4D1) (Santa Cruz, sc-2007)

snRNA-seq library preparation, sequencing, and data analysis

snRNA-seq libraries were prepared using the Split-seq platform (Butler et al., 2018; Rosenberg et al., 2018.). Freshly harvested mouse pons tissues underwent nuclear extraction following previously described protocols (Butler et al., 2018; Rosenberg et al., 2018.). In brief, mouse brain tissues were transferred into a 2 mL Dounce homogenizer containing 1mL homogenizing buffer (250 mM sucrose, 25 mM KCl, 5 mM MgCl2, and 10 mM Tris-HCl [pH = 8.0], 1 mM DTT, RNase Inhibitor and 0.1% Triton X-100). The mixture was subjected to five strokes with a loose pestle, followed by 10 strokes with a tight pestle. The resulting homogenates were filtered through a 40-μm strainer into 5-mL Eppendorf tubes and subsequently centrifuged for 4 min at 600 g and 4 °C. The pellet was re-suspended and washed in 1 mL of PBS containing RNase inhibitors and 0.1% BSA. The nuclei were again filtered through a 40-μm strainer and quantified. These nuclei were partitioned into 48 wells, each of which contained a barcoded, well-specified reverse transcription primer, to enable in-nuclear reverse transcription. Subsequent second and third barcoding steps were carried out through ligation reactions. After the third round of barcoding, the nuclei were divided into 16 aliquots and lysed prior to cDNA purification. The resulting purified cDNA was subjected to template switching and qRT-PCR amplification, which was halted at the beginning of the plateau stage. Finally, the purified PCR products (600 pg) were used to generate Illumina-compatible sequencing libraries. Distinct, indexed PCR primer pairs were employed to label each library, serving as the fourth barcode.

The libraries underwent sequencing on the NextSeq system (Illumina) using 150-nucleotide kits and paired-end sequencing protocols. In this arrangement, Read 1 covered the transcript sequences and Read 2 contained the UMI and UBC barcode combinations. Initially, a six-nucleotide sequencing index, serving as the fourth barcode, was appended to the ends of Read 2. Subsequently, reads with more than one mismatching base in the third barcode were excluded from the dataset. Furthermore, reads containing more than one low-quality base (Phred score ≤ 10) within the UMI region were also discarded. The sequencing results were aligned to exons and introns in the reference genome (Genome assembly GRCm38, https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000001635.20/) and aggregated intron and exon counts at the gene level were calculated by kallisto and bustools software as described (https://bustools.github.io/BUS_notebooks_R).

Following export of the matrix, quality control measures were performed to remove low-quality cells and potential doublets, as described previously (Rosenberg et al., 2018.). After filtering, a total of 125,956 and 122,266 cells for wild type and Rnf220+/−pons respectively remained for subsequent analysis. Seurat v2 (Butler et al., 2018.) was used to identify HVGs within each group. Principal component analysis (PCA) and UMAP were performed to embed each cell from the reduced principal component space on a 2D map. Then clustering of cell populations and identification of differentially expressed genes were carried out as previously described (Ma et al., 2022.). We annotated the embryonic cell populations and lineages based on their DEGs (Table 3).

Mouse brain stereotactic injection and neuronal tracing

Mouse brain stereotactic injection and neuronal tracing were carried out as previously described with minor modifications (Xu et al., 2021.). Adult mice (2–3 months old) were used for anterograde monosynaptic neuronal tracing. The mice were first deeply anesthetized with isoflurane and placed into a stereotactic apparatus with the front teeth latched onto the anterior clamp. The mouth and nose of the mice were covered with a mask to provide isoflurane and keep them in an anesthetic state during the operation. The head was adjusted and maintained in horizontal by inserting ear bars into the ear canal. The hair of the head was shaved with an electric razor and cleaned using a cotton swab which was dipped in 75% alcohol. Cut the scalp along the midline with surgical scissors and make sure the bregma and the interaural line were exposed. The intersection between the bregma and interaural line was set to zero. The coordinates (−1.75, −4.9, −1.6) were applied for motor cortex localization. Each mouse received a single injection of 1 µL of rAAV-hSyn-EGFP-WPRE-hGH-poly A (Brainvta, PT-1990) viral fluid in the left hemisphere. After the injection, mice were bred for another 3 weeks for neuronal tracing before their brains were harvested for analysis.

Slice preparation

Mice subjected to stereotactic injection were euthanized, and whole brains were collected for the preparation of frozen sections. Following fixation in 4% paraformaldehyde (diluted in PBS) for 48 h, the brains were successively treated with 20% sucrose (diluted in PBS) for 48 h and 30% sucrose for 24 h. Brain tissues were then embedded in Optimal Cutting Temperature (OTC) compound (SAKURA, 4583) at −20 ℃ and sliced along the sagittal plane at a thickness of 40 μm.

Quantification and statistical analysis

The fluorescent and chemiluminescent intensities were quantified by Image J software (National Institute of Health). Statistical analyses were performed using GraphPad Prism software (GraphPad Software Inc., La Jolla, CA, USA). Comparisons were performed using the two-tailed Student’s t-test. p-values of less than 0.05 were considered statistically significant (*), 0.01 were considered statistically significant (**) and 0.001 were considered statistically significant (***).

eLife Assessment

This useful study focuses on gene regulatory mechanisms essential for hindbrain development. Through molecular genetics and biochemistry, the authors propose a new mechanism for the control of Hox genes, which encode highly conserved transcription factors essential for hindbrain development. However, the strength of evidence is incomplete, as the main claims are only partially supported by the data. This work will be of interest to developmental biologists.

Reviewer #1 (Public Review):

The manuscript by Wang et al. investigates the role of Rnf220 in hindbrain development and Hox expression. The authors suggest that Rnf220 controls Hox expression in the hindbrain by regulating WDR5 levels. The authors combine in vivo experiments with experiments in P19 cells to demonstrate this mechanism. However, the in vivo data does not provide strong support for the claims the authors make and the role of Rnf in Hox maintenance and pons development is unclear.

Specific concerns with in vivo data:

A major issue throughout the paper is that Hox expression analysis is done exclusively through quantitative PCR, with values ranging from 2-fold to several thousand-fold upregulation, with no antibody validation for any Hox protein (presumably they are all upregulated).

In Figure 1, massive upregulation of most Hox genes in the brainstem is shown after e16.5 but the paper quickly focuses on analysis of PN nuclei. What are the other consequences of this broad upregulation of Hox genes in the brainstem? There is no discussion of the overall phenotype of the mice, the structure of the brainstem, the migration of neurons, etc. The very narrow focus on motor cortex projections to PN nuclei seems bizarre without broad characterization of the mice, and the brainstem in particular. There is only a mention of "severe motor deficits" from previous studies, but given the broad expression of Rnf220, the fact that is a global knockout, and the effects on spinal cord populations shown previously the justification for focusing on PN nuclei does not seem strong.

It is stated that cluster 7 in scRNA-seq corresponds to the PN nuclei. The modest effect shown on Hox3-5 expression in that data in Figure 1 is inconsistent with the larger effect shown in Figure 2.

Presumably, Hox genes are not the only targets of Rnf220 as shown in the microarray/RNA-sequencing data. There is no definitive evidence that any phenotypes observed (which are also not clear) are specifically due to Hox upregulation. The only assay the authors use to look at a Hox-dependent phenotype in the brainstem is the targeting of PN nuclei by motor cortex axons. This is only done in 2 animals and there are no details as to how the data was analyzed and quantified. The only 2 images shown are not convincing of a strong phenotype, they could be taken at slightly different levels or angles. At the very least, serial sections should be shown and the experiment repeated in more animals. There is also no discussion of how these phenotypes, if real, would relate to previous work by the Rijli group which showed very precise mechanisms of synaptic specificity in this system.

The temporal aspect of this regulation in vivo is not clear. The authors show some expression changes begin at e16.5 but are also present at 2 months. Is the presumed effect on neural circuits a result of developmental upregulation at late embryonic stages or does the continuous overexpression in adult mice have additional influence? Are any of the Hox genes upregulated normally expressed in the brainstem, or PN specifically, at 2 months? Why perform single-cell sequencing experiments at 2 months if this is thought to be mostly a developmental effect? Similarly, the significance of the upregulated WRD5 in the pons and pontine nuclei at 2 months in Figure 3 is not clear.

In Figure 3C the levels of RNF220 in wt and het don't seem to be that different.

Based on the single-cell experiments, and the PN nuclei focus, the rescue experiments are confusing. If the Rnf220 deletion has a sustained effect for up to 2 months, why do the injections in utero? If the focus is the PN nuclei why look at Hox9 expression and not Hox3-5 which are the only Hox genes upregulated in PN based on sc-sequencing? No rescue of behavior or any phenotype other than Hox expression by qPCR is shown and it is unclear whether upregulation of Hox9 paralogs leads to any defects in the first place. The switch to the Nes-cre driver is not explained. Also, it seems that wdr5 mRNA levels are not so relevant and protein levels should be shown instead (same for rescue experiments in P19 cells).

Other:

What is the relationship between Retinoic acid and WRD5? In Figure 3E there is no change in WRD5 levels without RA treatment in Rnf KO but an increase in expression with RA treatment and Rnf KO. However, the levels of WRD5 do not seem to change with RA treatment alone. Does Rnf220 only mediate WDR5 degradation in the presence of RA? This does not seem to be the case in experiments in 293 cells in Figure 4.

Why are the levels of Hox upregulation after RA treatment so different in Figure 5 and Figure Supplement 5?

In Figures 4B+C which lanes are input and which are IP? There is no quantitation of Figure 4D, from the blot it does look that there is a reduction in the last 2 columns as well. The band in the WT flag lane seems to have a bubble. Need to quantitate band intensities. Same for E, the effect does not seem to be completely reversed with MG132.

Reviewer #2 (Public Review):

Wang, Liu, et al. identified Rnf220 and Wdr5 as novel regulators of Hox gene expression during pons development. Phenotypic characterization of Rnf220 deficient mice with single-cell transcriptomics, qRT-PCR, and axonal tracing methods show that Rnf220 knockdown causes de-repression of Hox gene expression at multiple stages of pons development to regulate the final formation of the pontine nuclei neural circuit. Additionally, they also perform exhaustive expression analysis of multiple genes in the Hox family cluster to identify specific gene groups that are targeted by Rnf220. Furthermore, they also demonstrate that Rnf220 modulates Hox gene expression by directly binding to Wdr5, thus targeting it for ubiquitination and subsequent degradation. To elucidate the molecular mechanism of this interaction, they perform detailed immunoprecipitation assays and identify the precise Wdr5 amino acid residues that are targeted by Rnf220. Intriguingly, they show that inhibition of Wdr5 in Rnf220 deficient mice reverses the de-repression of Hox gene expression suggesting the direct involvement of Rnf220-Wdr5 interaction in modulating Hox gene expression during pons development. These data highlight the role of a new form of Hox gene regulation via the ubiquitination of epigenetic modulator Wdr5.

The conclusions of this paper are mostly supported by the data provided, but the downstream molecular and tissue-level effects of Wdr5 knockdown/inhibition need to be further characterized to establish its definitive role in pons development.

(1) Figure 1E shows that Rnf220 knockdown alone could not induce an increase in Hox expression without RA, which indicates that Rnf220 might endogenously upregulate Retinoic acid signaling. The authors should test if RA signaling is downstream of Rnf220 by looking at differences in the expression of Retinaldehyde dehydrogenase genes (as a proxy for RA synthesis) upon Rnf220 knockdown.

(2) In Figure 2C-D further explanation is required to describe what criteria were used to segment the tissue into Rostral, middle, and caudal regions. Additionally, it is unclear whether the observed change in axonal projection pattern is caused due to physical deformation and rearrangement of the entire Pons tissue or due to disruption of Hox3-5 expression levels. Labeling of the tissue with DAPI or brightfield image to show the structural differences and similarities between the brain regions of WT and Rnf220 +/- will be helpful.

(3) Line 192-195. These roles of PcG and trxG complexes are inconsistent with their initial descriptions in the text - lines 73-74.

(4) In Figure 4D, the band in the gel seems unclear and erased. Please provide a different one. These data show that neither Rnf220 nor wdr5 directly regulates Hox gene expressions. The effect of double knockdown in the presence of RA suggests that they work together to suppress Hox gene expression via a different downstream target. This point should be addressed in the text and discussion section of the paper. example for the same data which shows a full band with lower intensity.

(5) In Figure 4G the authors could provide some form of quantitation for changes in ubiquitination levels to make it easier for the reader. They should also describe the experimental procedures and conditions used for each of the pull-down and ubiquitination assays in greater detail in the methods section.

(6) Figure 5 shows that neither Rnf220 nor wdr5 directly regulate Hox gene expressions. The effect of double knockdown in the presence of RA suggests that they work together to suppress Hox gene expression via a different downstream target. This point should be addressed in the text and discussion section of the paper.

(7) In Figure 6, while the reversal of changes in Hox gene expression upon concurrent Rnf220; Wdr5 inhibition highlights the importance of Wdr5 in this regulatory process, the mechanistic role of wdr5 and its functional consequences are unclear. To answer these questions, the authors need to: (i) Assay for activated and repressive epigenetic modifications upon double knockdown of Rnf220 and Wdr5 similar to that shown in Figure 4- supplement 1. This will reveal if wdr5 functions according to its intended role as part of the TrxG complex. (ii) The authors need to assay for changes in axon projection patterns in the double knockdown condition to see if Wdr5 inhibition rescues the neural circuit defects in Rnf220 +/- mice.

Author response:

The following is the authors’ response to the original reviews.

Reviewer 1:

(1) A major issue throughout the paper is that Hox expression analysis is done exclusively through quantitative PCR, with values ranging from 2-fold to several thousand-fold upregulation, with no antibody validation for any Hox protein (presumably they are all upregulated).

Thank you for your comment.

We tried to verify the stimulated Hox expression pattern by in situ hybridization. Although in early embryos (E9.5) we could detect clearly hox (i.e. Hox8 and Hox9 in Author response image 1) expression patterns in the neural tube by whole mount in situ hybridization, we failed to detect a clear pattern in the brain stem at E18.5 either in whole mount tissue or on sections. That’s one reason that we turned to single nuclear RNA-seq instead.

This is likely due to their low expression levels at late developmental stages and need to be detected by more sensitive method. However, we estimated that the stimulated expression levels of the representative Hox genes are at least comparable to the physiological levels at posterior spinal cord to evoke a functional effect.

Some Hox8 and Hox9 expression pattern in E9.

5 embryos.

(2) In Figure 1, massive upregulation of most Hox genes in the brainstem is shown after e16.5 but the paper quickly focuses on analysis of PN nuclei. What are the other consequences of this broad upregulation of Hox genes in the brainstem? There is no discussion of the overall phenotype of the mice, the structure of the brainstem, the migration of neurons, etc. The very narrow focus on motor cortex projections to PN nuclei seems bizarre without broad characterization of the mice, and the brainstem in particular. There is only a mention of "severe motor deficits" from previous studies, but given the broad expression of Rnf220, the fact that is a global knockout, and the effects on spinal cord populations shown previously the justification for focusing on PN nuclei does not seem strong.

Thank you for your comment.

Although RNF220 is important for the dorsal-ventral patterning of the spinal cord as well as the hindbrain during embryonic development, the earlier neural patterning and differentiation are normal in the Rnf220+/- mice (Wang et al., 2022). However, these mice showed reduced survival and motility to various degree postnatally (Ma et al., 2019; Ma et al., 2021), likely suggesting a dosage dependent role of RNF220 in maintaining late neural development. As our microarray assay showed the deregulation of the Hox genes in the brain, we followed this direction in this study and narrowed down the affected region to the pons. Our single nuclear RNA-Seq (snRNA-seq) data further shows that the Hox de-regulation mainly occurred in 3 clusters of neurons. However, the pons is complex and contains tens of nuclei. And the current resolution of our data does not support to assign a clear identity to each of them. Although it is clear that more nuclei are likely affected, the PN (cluster7) is the only cluster we can identify to follow in the current study.

As to general effect of RNF220 haploinsufficiency on the brainstem, we carried out Nissl staining assays and found no clear difference in neuronal cell organization between WT and Rnf220+/- pons (revised Figure 2-figure supplement 2).

(3) It is stated that cluster 7 in scRNA-seq corresponds to the PN nuclei. The modest effect shown on Hox3-5 expression in that data in Figure 1 is inconsistent with the larger effect shown in Figure 2.

Thank you for your comment.

Due to the low efficiency of snRNA-seq and the depth of the sequencing, the quantification of the Hox expression based on the snRNA-seq data is likely less accurate as the qRT-PCR. In addition, only mRNAs in the nuclear could be captured by snRNA-seq, while mRNAs in both the nuclear and cytoplasm were reversed-transcribed and examined for qRT-PCR assays in Figure 2A.

(4) Presumably, Hox genes are not the only targets of Rnf220 as shown in the microarray/RNA-sequencing data. There is no definitive evidence that any phenotypes observed (which are also not clear) are specifically due to Hox upregulation. The only assay the authors use to look at a Hox-dependent phenotype in the brainstem is the targeting of PN nuclei by motor cortex axons. This is only done in 2 animals and there are no details as to how the data was analyzed and quantified. The only 2 images shown are not convincing of a strong phenotype, they could be taken at slightly different levels or angles. At the very least, serial sections should be shown and the experiment repeated in more animals. There is also no discussion of how these phenotypes, if real, would relate to previous work by the Rijli group which showed very precise mechanisms of synaptic specificity in this system.

Thank you for your comments and suggestions.

The deregulation of Hox is the most obvious phenomena observed from the RNA-seq data, and we tried to assign its specific phenotypic effect in this study. As the roles of Hox in PN patterning and circuit formation is well established, we focused on the PN in the following study. Based on literature, we carried out the circuit analysis to examine the targeting of PN neurons by the motor cortex axons. A cohort of additional animals with different genotypes (n=10 for WT and n=9 for Rnf220+/-) were used to repeat the experiment and we got the same conclusion. More detailed information on data analysis and serial images were included in the revised manuscript and figure legends.

(5) The temporal aspect of this regulation in vivo is not clear. The authors show some expression changes begin at e16.5 but are also present at 2 months. Is the presumed effect on neural circuits a result of developmental upregulation at late embryonic stages or does the continuous overexpression in adult mice have additional influence? Are any of the Hox genes upregulated normally expressed in the brainstem, or PN specifically, at 2 months? Why perform single-cell sequencing experiments at 2 months if this is thought to be mostly a developmental effect? Similarly, the significance of the upregulated WRD5 in the pons and pontine nuclei at 2 months in Figure 3 is not clear.

Thank you for your comment.

The spatial and temporal expression pattern of Hox genes is established at early embryonic stages and then maintained throughout developmental stage in mammals. As we have shown, the de-repression of Hox genes is a long-lasting defect in Rnf220+/- mice beginning at late embryonic stages. Since the neuronal circuit is established after birth in mice, we speculated that the neuronal circuit defects from motor cortex to PN neurons were due to the long-lasting up-regulation of Hox genes in PN neurons. We could not distinguish the effect on neural circuit a result of Hox genes developmental upregulation or continuous overexpression in adult mice. An inducible knockout mouse model may help to answer this question in the future. The discussion on this point was included in the revised manuscript.

We carried out snRNA-seq analysis using pons tissues from adult mice aiming to identify the specific cell population with Hox up-regulation, which we failed to specify by in situ hybridization.

We repeated the related experiments in the original Figure 3 and some of the blot images were replaced and quantified.

(6) In Figure 3C, the levels of RNF220 in wt and het don't seem to be that different.

We repeated the experiments and changed the related image in the revised Figure 3C.

(7) Based on the single-cell experiments, and the PN nuclei focus, the rescue experiments are confusing. If the Rnf220 deletion has a sustained effect for up to 2 months, why do the injections in utero? If the focus is the PN nuclei why look at Hox9 expression and not Hox3-5 which are the only Hox genes upregulated in PN based on sc-sequencing? No rescue of behavior or any phenotype other than Hox expression by qPCR is shown and it is unclear whether upregulation of Hox9 paralogs leads to any defects in the first place. The switch to the Nes-cre driver is not explained. Also, it seems that wdr5 mRNA levels are not so relevant and protein levels should be shown instead (same for rescue experiments in P19 cells).

Thank you for your comments.

Since our data suggest that the upregulation of Hox genes expression is a long-lasting effect beginning at the late embryonic stage of E16.5, we conducted the rescue experiments by in utero injection of WDR5 inhibitor at E15.5 and examined the expression of Hox genes at E18.5. Although it is also necessary to examine whether the rescue effect by WDR5 inhibitor injection is also a long-lasting effect at adult stages, it is difficult to distinguish the embryos or pups when they were given birth. As a supplement, rescue assays with genetic ablation of Wdr5 gene were conducted and the results showed that genetic ablation of a single copy of Wdr5 allele could revere the upregulation of Hox genes by RNF220 haploinsufficiency in the hindbrains at P15.

Most of the upregulated Hox genes including both Hox9 and Hox3-5 were examined in our rescue experiments. Since this study focuses on the PN nuclei, the results of Hox3-5 genes were shown in the revised main Figure 6.

We conducted rescue experiments by deleting Wdr5 in neural tissue using _Nestin-Cr_e mice because Wdr5+/- mice is embryonic lethal. And the up-regulation of Hox genes could be also observed in the hindbrains of Rnf220fl/wt; Nestin-Cre mice. Although Rnf220fl/wt; Wdr5fl/wt; Nestin-Cre mice are viable and could survive to adult stages, developmental defects in the forebrains, including cerebral cortex and hippocampus, were observed in Rnf220fl/wt;Wdr5fl/wt;Nestin-Cre mice. Therefore, no rescue of behavior tests was conducted in this study. We believe that it is out of the scope of this study to discuss the role of WDR5 in the development of forebrains.

The potential defects due to the up-regulation of Hox9 paralogs awaits further investigations.

Wdr5 mRNA levels were firstly examined to confirm the genetic deletion or siRNA mediated knockdown of Wdr5 genes. We have carried out western blot to examine the WDR5 protein levels and the results were included in the revised Figure 3.

(8) What is the relationship between Retinoic acid and WRD5? In Figure 3E there is no change in WRD5 levels without RA treatment in Rnf KO but an increase in expression with RA treatment and Rnf KO. However, the levels of WRD5 do not seem to change with RA treatment alone. Does Rnf220 only mediate WDR5 degradation in the presence of RA? This does not seem to be the case in experiments in 293 cells in Figure 4.

Thank you for your comment.

We believe that the regulation of WDR5 and Hox expression by RNF220 is context dependent and precisely controlled in vivo, depending on the molecular and epigenetic status of the cell, which is fulfilled by RA treatment in P19 cells. In Figure 4, the experiment is based on exogenous overexpression assays, which might not fully reflect the situation in vivo.

(9) Why are the levels of Hox upregulation after RA treatment so different in Figure 5 and Figure Supplement 5?

In Figure.5C, the Hox expression levels were normalized against the control group in the presence of RA; while in Figure Supplement 5 they were normalized to the control group without RA treatment.

(10) In Figures 4B+C which lanes are input and which are IP? There is no quantitation of Figure 4D, from the blot it does look that there is a reduction in the last 2 columns as well. The band in the WT flag lane seems to have a bubble. Need to quantitate band intensities. Same for E, the effect does not seem to be completely reversed with MG132.

Thanks for pointing this out. The labels were included in the revised Figure 4B and 4C.

We repeated the experiments for Figure 4D and 4E. Some of bot images were replaced and quantified in the revised Figure 4D and 4E.

Reviewer 2:

(1) Figure 1E shows that Rnf220 knockdown alone could not induce an increase in Hox expression without RA, which indicates that Rnf220 might endogenously upregulate Retinoic acid signaling. The authors should test if RA signaling is downstream of Rnf220 by looking at differences in the expression of Retinaldehyde dehydrogenase genes (as a proxy for RA synthesis) upon Rnf220 knockdown.

Thank you for your comment and suggestion.

Two sequential reactions are required for RA synthesis from retinol, which catalyzed by alcohol dehydrogenases (ADHs)/ retinol dehydrogenase (RDH) and retinaldehyde dehydrogenase (RALDHs also known as ALDHs) respectively. When RA is no longer needed, it is catabolized by cytochrome enzymes (CYP26 enzymes) (Niederreither, et al.,2008; Kedishvili et al., 2016). Here, we test ADHs、ALDHs and CYP26 enzymes in E16.5 WT and Rnf220-/- embryos.

The results are as follows. ADH7 and ADH10 are slightly upregulated. ALDH1 and ALDH3 are upregulated and downregulated in Rnf220-/- embryos, respectively, but there is no significant change in the expression of ALDH2, which plays a key role in RA synthesis during embryonic development (Niederreither, et al.,2008). Furthermore, Cyp26a1 which responsible for RA catabolism was upregulated in Rnf220-/- embryos. Collectively, these data do not support a clear effect on RA signaling by RNF220.

The effect of Rnf220 on RA synthesis and degradation pathways

(2) In Figure 2C-D further explanation is required to describe what criteria were used to segment the tissue into Rostral, middle, and caudal regions. Additionally, it is unclear whether the observed change in axonal projection pattern is caused due to physical deformation and rearrangement of the entire Pons tissue or due to disruption of Hox3-5 expression levels. Labeling of the tissue with DAPI or brightfield image to show the structural differences and similarities between the brain regions of WT and Rnf220 +/- will be helpful.

Thank you for your comment and suggestion.

More information on the quantification of the results shown in Figure 2C-D was included in our revised manuscript. We carried out Nissl staining assays using coronal sections of the brainstem and found that there is no significant difference in neuronal cell organization between WT and Rnf220+/- (revised Figure 2-figure supplement 2).

(3) Line 192-195. These roles of PcG and trxG complexes are inconsistent with their initial descriptions in the text - lines 73-74.

We are sorry for the mistake. We carefully revised the related descriptions to avoid such mistake. Thank you.

(4) In Figure 4D, the band in the gel seems unclear and erased. Please provide a different one. These data show that neither Rnf220 nor wdr5 directly regulates Hox gene expressions. The effect of double knockdown in the presence of RA suggests that they work together to suppress Hox gene expression via a different downstream target. This point should be addressed in the text and discussion section of the paper. example for the same data which shows a full band with lower intensity.

Thank you for your suggestion.

We repeated the experiment of Figure 4D and some of the blot images were replaced in the revised Figure 4D.

Indeed, in the presence of RA, knockdown of Rnf220 alone can upregulate the expression Hox genes (Figure 5C). Knockdown of Wdr5 could reverse the upregulation of Hox genes in RNF220 knockdown cells, suggesting that Rnf220 regulated Hox gene expression in a Wdr5 dependent manner. However, in the absence of RA, none of Rnf220 knockdown, Wdr5 knockdown or Rnf220 and Wdr5 double knockdown had a significant effect on the expression of Hox genes in P19 cells. It seems that RA signaling plays a crucial role for the regulation of RNF220 to WDR5 in P19 cells and discussion on this point was included in the revised manuscript.

(5) In Figure 4G the authors could provide some form of quantitation for changes in ubiquitination levels to make it easier for the reader. They should also describe the experimental procedures and conditions used for each of the pull-down and ubiquitination assays in greater detail in the methods section.

Thank you for your suggestion.

The quantitation and statistics for the original Figure 4G were included in the revised Figure 4. More information on the biochemical assays was included in the “Methods and Materials” section of our revised manuscript.

(6) Figure 5 shows that neither Rnf220 nor wdr5 directly regulate Hox gene expressions. The effect of double knockdown in the presence of RA suggests that they work together to suppress Hox gene expression via a different downstream target.

Thank you for your comment.

In fact, knockdown of Rnf220 alone can upregulate the expression Hox genes in the presence of RA (Figure 5C). Furthermore, knockdown of Wdr5 could reverse the upregulation of Hox genes in Rnf220 knockdown cells, which suggest that Rnf220 regulated Hox gene expression in a Wdr5 dependent manner. However, in the absence of RA, none of Rnf220 knockdown, Wdr5 knockdown or Rnf220 and Wdr5 double knockdown had a significant effect on the expression of Hox genes in P19 cells. It seems that RA signaling plays a crucial role for the regulation of RNF220 to WDR5 in P19 cells and discussion on this point was included in the revised manuscript.

(7) In Figure 6, while the reversal of changes in Hox gene expression upon concurrent Rnf220; Wdr5 inhibition highlights the importance of Wdr5 in this regulatory process, the mechanistic role of wdr5 and its functional consequences are unclear. To answer these questions, the authors need to: (i) Assay for activated and repressive epigenetic modifications upon double knockdown of Rnf220 and Wdr5 similar to that shown in Figure 3- supplement 1. This will reveal if wdr5 functions according to its intended role as part of the TrxG complex. (ii) The authors need to assay for changes in axon projection patterns in the double knockdown condition to see if Wdr5 inhibition rescues the neural circuit defects in Rnf220 +/- mice.

Thank you for your suggestion.

Although it is also necessary to examine whether the rescue effect by WDR5 inhibitor injection in uetro is also a long-lasting effect for neuronal cirtuit at adult stages, it is difficult to distinguish the embryos or pups when they were given birth. Although Rnf220fl/wt;Wdr5fl/wt;Nestin-Cre mice are viable and could survive to adult stages, developmental defects in the forebrains, including cerebral cortex and hippocampus, were observed in Rnf220fl/wt;Wdr5fl/wt;Nestin-Cre mice. Therefore, no rescue effect on defects of behavior and neuronal circuit were examined in this study. Maybe, a PN nuclei specific inducible Cre mouse line could help toward this direction in the future.

We carried out ChIP-qPCR and tested activated and repressive epigenetic modifications upon double knockdown of Rnf220 and Wdr5 in P19 cell line and found Rnf220 and Wdr5 double knockdown recured Hox epigenetic modification to a certain degree (Figure 6-figure supplement 1).

References

Kedishvili, N.Y. 2016. Retinoic acid synthesis and degradation. Subcell Biochem, 81:127-161. DOI: 10.1007/978-94-024-0945-1_5, PMID: 2783050

Ma, P., Li, Y., Wang, H., Mao, B., Luo, Z.-G. 2021. Haploinsufficiency of the TDP43 ubiquitin E3 ligase RNF220 leads to ALS-like motor neuron defects in the mouse. Journal of Molecular Cell Biology, 13: 374-382. DOI: 10.1093/jmcb/mjaa072, PMID: 33386850

Ma, P., Song, N.-N., Li, Y., Zhang, Q., Zhang, L., Zhang, L., Kong, Q., Ma, L., Yang, X., Ren, B., Li, C., Zhao, X., Li, Y., Xu, Y., Gao, X., Ding, Y.-Q., Mao, B. 2019. Fine-Tuning of Shh/Gli Signaling Gradient by Non-proteolytic Ubiquitination during Neural Patterning. Cell Rep, 28: 541-553.e544. DOI: 10.1016/j.celrep.2019.06.017, PMID: 31291587

Niederreither, K., Dollé, P. 2008. Retinoic acid in development: towards an integrated view. Nat Rev Genet, 9: 541-53. DOI: 10.1038/nrg2340, PMID: 18542081

Wang, Y.-B., Song, N.-N., Zhang, L., Ma, P., Chen, J.-Y., Huang, Y., Hu, L., Mao, B., Ding, Y.-Q. 2022. Rnf220 is Implicated in the Dorsoventral Patterning of the Hindbrain Neural Tube in Mice. Front Cell Dev Biol, 10. DOI: 10.3389/fcell.2022.831365, PMID: 35399523

Version 2

Abstract

Abstract

The spatial and temporal linear expression of Hox genes establishes a regional Hox code, which is crucial for the antero-posterior (A-P) patterning, segmentation, and neuronal circuit development of the hindbrain. RNF220, an E3 ubiquitin ligase, is widely involved in neural development via targeting of multiple substrates. Here, we found that the expression of Hox genes in the pons was markedly up-regulated at the late developmental stage (post-embryonic day E15.5) in Rnf220−/− and Rnf220+/−mouse embryos. Single-nucleus RNA-seq analysis revealed different Hox de-repression profiles in different groups of neurons, including the pontine nuclei (PN). The Hox pattern was disrupted and the neural circuits were affected in the PN of Rnf220+/− mice. We showed that this phenomenon was mediated by WDR5, a key component of the TrxG complex, which can be polyubiquitinated and degraded by RNF220. Intrauterine injection of WDR5 inhibitor (WDR5-IN-4) and genetic ablation of Wdr5 in Rnf220+/− mice largely recovered the de-repressed Hox expression pattern in the hindbrain. In P19 embryonal carcinoma cells, the retinoic acid induced Hox expression was further stimulated by Rnf220 knockdown, which can also be rescued by Wdr5 knockdown. In short, our data suggest a new role of RNF220/WDR5 in Hox pattern maintenance and pons development in mice.

Introduction

The coordinated expression of Hox genes clusters is critical for axial patterning to determine the positional identities during the development of many tissues, including the hindbrain (Frank et al., 2019; Parker et al., 2014; Parker et al., 2020; Pöpperl et al., 1993). In mammals, there are 39 Hox genes organized in four clusters (HoxA, HoxB, HoxC, and HoxD) and divided into 1-13 paralog groups (PG). During vertebrate development, the hindbrain forms eight metameric segmented units along the antero-posterior (A-P) axis known as rhombomeres (r), which give rise to the cerebellum, pons and medulla later on (Ghosh et al., 2018; Parker et al., 2020). In the embryonic hindbrain, Hox1-5 genes are expressed in a nested overlapping pattern with rhombomere-specific boundaries. This Hox code is critical for the establishment of rhombomeric territories and the determination of cell fates (Barsh et al., 2017; Parker et al., 2016; Parker et al., 2020).

At later embryonic stages, Hox genes also play vital roles in neuronal migration and circuit formation (Feng et al., 2021; Hockman et al., 2019; Kratochwil et al., 2017; Wang et al., 2013). For example, neurons in the pontine nuclei (PN), which acts as an information integration station between the cortex and cerebellum (Kratochwil et al., 2017; Maheshwari et al., 2020), originate at the posterior rhombic lip of r6-r8, migrate tangentially to the rostral ventral hindbrain, and then receive different inputs from the cortex, in which the specific expression pattern of Hox3-5 genes determines the migration routines, cellular organization, and neuronal circuits (Geisen et al., 2008; Kratochwil et al., 2017; Maheshwari et al., 2020; Di Meglio et al., 2013; Lizen et al., 2017). Note that Hox gene expression is maintained up to postnatal and even adult stages in the hindbrain derivatives (Farago et al., 2006; Feng et al., 2021; Kratochwil et al., 2017).

Hox pattern maintenance is generally governed by epigenetic regulators, especially the polycomb-group (PcG) and trithorax-group (trxG) complexes, which repress and activate Hox genes, respectively (Bahrampour et al., 2019; Chopra et al., 2009; Kang et al., 2022; Papp et al., 2006). In mammals, different SET/MLL proteins, and a common multi-subunit core module consisting of WDR5, RbBP5, Ash2L, and DPY-30 (WRAD) constitute the COMPASS (complex of proteins associated with Set1) family of TrxG complexes, which acts as a methyltransferase generally (Cenik et al., 2021; Jambhekar et al., 2019; Schuettengruber et al., 2017). Among them, WDR5 plays a central scaffolding role and has been shown to act as a key regulator of Hox maintenance (Wang et al., 2011; Wysocka et al., 2005). In acute myeloid leukemia (AML) and acute lymphoblastic leukemia (ALL), WDR5 is responsible for MLL-mediated HoxA9 activation and is a potential drug target (Chen et al., 2021; Yu et al., 2021). In addition, WDR5 facilitates HOTTIP, a long noncoding RNA (lncRNA), to recruit the MLL complex and thus stimulate HoxA9 and HoxA13 expression during the progression of prostate and liver tumors (Fu et al., 2017; Malek et al., 2017; Quagliata et al., 2014; Wong et al., 2020).

The E3 ubiquitin ligase RNF220 is widely involved in neural development through targeting a range of proteins (Kim et al., 2018; Kong et al., 2010; Ma et al., 2021; Ma et al., 2022; Wang et al., 2022). Although our previous reports have shown that Rnf220−/− mice is neonatal lethal and the survived Rnf220+/− mice develop severe motor impairments; the underlying mechanisms remain incompletely understood (Ma et al., 2021; Ma et al., 2019). In the present study, a markedly up-regulation of Hox genes is observed in the pons of both Rnf220+/− and Rnf220−/− mice during late embryonic development. In consequence, the topographic input connectivity of the PN neurons with cortex is disturbed in Rnf220+/− mice. Furthermore, we identified WDR5 as a direct target of RNF220 for K48-linked ubiquitination and thus degradation, a mechanism critical for Hox de-repression by Rnf220 knockout. Together, these findings reveal a novel role for RNF220 in the regulation of Hox genes and pons development in mice.

Results

Rnf220 insufficiency leads to dysregulation of Hox expression in late embryonic pons

Rnf220 is strongly expressed in the embryonic mouse mid/hind brain (Ma et al., 2019; Wang et al., 2022). Based on microarray analysis, we explored changes in expression profiles in E18.5 Rnf220+/− and Rnf220−/− mouse brains, which breed floxed Rnf220 to Vasa-Cre mice. Interestingly, several Hox genes were up-regulated in the mouse brain of both genotypes (Tables 1 and 2). Considering that Hox expression in forebrain is too low (data not shown), we focus on the hindbrain and found that this phenomenon was observed only in late embryonic stages after E15.5 (Figure 1-figure supplement 1). Since Rnf220−/− mice is neonatal lethal and there is not significant diference in the upregulation of Hox genes between Rnf220+/− and Rnf220−/− mice, Rnf220+/− mice were used in subsequent assays in this study. To determine the specific brain regions with aberrant Hox expression, we examined the expression levels of HoxA9 and HoxB9, two of the most significantly up-regulated genes, in different brain and spinal regions (Figure 1-figure supplement 2A-D). Notably, both HoxA9 and HoxB9 were exclusively up-regulated in the brainstem of Rnf220+/− mice (Figure 1-figure supplement 2C and D). The brainstem comprises the midbrain, pons, and medulla (Figure 1-figure supplement 2E). Further, we confirmed that the upregulation of Hox genes was restricted to the pons in the Rnf220+/− mice (Figure 1-figure supplement 2F). In addition, RNA sequencing (RNA-seq) analysis of the pons also revealed an overall increase in Hox gene expression in adult Rnf220+/− mice (Figure 1A).

Hox genes up-regulated in pons of Rnf220+/− mice.

(A) The heatmap of RNA-seq data showing Hox genes expression in pons of WT or Rnf220+/− mice (n=2 mice per group). (B-C) UMAP diagram showing 15 identified cell clusters annotated by snRNA-seq analysis of pons. Each dot represents a single cell, and cells are laid out to show similarities (n=3 mice per group). (D) Heatmap of snRNA-seq data showing Hox expression changes in each cell cluster. (E) Quantitative real time polymerase chain reaction (qRT-PCR) analysis showing mRNA levels of indicated Hox genes in P19 cells when endogenous Rnf220 was knocked down by siRNAs in the presence of RA. WT, wild-type. HE, heterozygote. RA, retinoic acid. **p < 0.01, ***p < 0.001.

To identify the specific cell populations in which Hox genes were up-regulated, we conducted single-nucleus RNA-seq (snRNA-seq) analysis of the pons of adult wild-type (WT) and Rnf220+/− mice. In total, 125956 and 122266 cell transcriptomes with an average of approximately ≥700 genes for the WT and Rnf220+/−mice respectively were further analyzed. 15 cell clusters were identified by uniform manifold approximation and projection (UMAP) in both the WT and Rnf220+/− groups (Figure 1B and C). When we annotate these cell clusters by their uniquely and highly expressed markers (Table 3), we found that most of the clusters were identified as distinct neuronal groups (cluster_1, 2, 4, 6, 7, 8, 10, 11, 12, and 13), in addition to astrocyte (cluster_5), oligodendrocyte (cluster_3), oligodendrocyte precursor (cluster_14), ependymal cell (cluster_9), and a group not corresponding to any known cell types (cluster_0) (Figure 1B). Then the Hox gene expression levels for each cell clusters were analyzed and found that up-regulation of Hox genes was most pronounced in three of these clusters (cluster_2, 7, and 10) (Figure 1D), albeit with distinct profiles. Hox7–10 genes were up-regulated genes in cluster_2, while Hox3–5 and Hox1–3 genes were activated in cluster_7 and 10, respectively (Figure 1D). These findings indicate that Hox genes were specifically up-regulated in the pons of Rnf220+/− mice from late developmental stages on.

P19 embryonal carcinoma cells can be induced to differentiate and express Hox genes upon retinoic acid (RA) administration and have been used to study Hox regulation (Heyden et al., 2003; Kondo et al., 1992; Pöpperl et al., 1993; Figure 1-figure supplement 3A). Here, we tested the effects of Rnf220 knockdown on RA-induced Hox gene activation in P19 cells. Rnf220 knockdown further enhanced the activation of many Hox genes, including HoxA1, HoxB1, HoxA9, and HoxB9, by RA, (Figure 1E), suggesting a general role of Rnf220 in Hox regulation. Note that Rnf220 knockdown had no clear effect on Hox expression in the absence of RA (Figure 1-figure supplement 3B).

Rnf220 insufficiency disturbs Hox expression pattern and the neuronal circuit formation in the PN

When we refer to the transcriptomic profiles of each cluster in the pons, we found that the cluster_10 likely represents cells derived from rhombomeres 2–5, given their expression of endogenous anterior Hox genes (Hox1–3), while the cluster_7 likely represents cells derived from rhombomeres 6–8, as evidenced by the expression of Hox3–5 (Figure 1D). After mapping known neuron-specific markers to each cluster, PN-specific markers, including Nfib, Pax6, and Barhl, were enriched in the cluster_7 (Figure 2-figure supplement 1A and B). Originating at the posterior rhombic lip and migrating tangentially to their final position in the ventral part of the pons, PN neurons serve as relay cells between the cerebral cortex and cerebellum (Di Meglio et al., 2013; Kratochwil et al., 2017; Maheshwari et al., 2020). The expression pattern of Hox genes is crucial for the migration and the final neuronal circuit formation of the PN neurons (Kratochwil et al., 2017; Maheshwari et al., 2020). We analyzed the expression pattern of the endogenous Hox genes, including Hox3-5, in the PN neurons and the connection between the cortex and the PN neurons in Rnf220+/− mice. The results of realtime RT-PCR assays indicated an up-regulation in Hox4–5 expression levels in the Rnf220+/− PN (Figure 2-figure supplement 1C). In the PN, endogenous Hox3–5 genes exhibit a nested and unique expression pattern along the rostral-caudal axis, with Hox3 ubiquitously expressed throughout the PN, Hox4 localized to the middle and posterior regions, and Hox5 confined to the posterior segment (Kratochwil et al., 2017). Nest, the PN was evenly dissected into the rostral, middle, and caudal segments alone the rostral-caudal axis and the endogenous expression pattern of Hox genes was examined in the mice. The results showed that all the Hox3-5 paralogs were uniformly up-regulated along the rostral-caudal axis in in the Rnf220+/− PN (Figure 2A).

Hox gene expression was dysregulated and motor cortex projections were disorganized in PN of Rnf220+/− mice.

(A) qRT-PCR analysis of relative expression levels of Hox3, Hox4, and Hox5 in rostral, middle, and caudal sections of PN in WT and RNF220+/− mice. Expression level of each gene in rostral section of WT PN was set to 1. (n=5 mice per group). Bar graphs show the relative levels normalized against rostral group in the respective wild-type mice. (B) Diagram of experimental stereotactic injections. (C) Green fluoresce showed the projection from motor cortex to PNs in adult (2 months) WT (n=10 mice) and Rnf220+/− mice (n=9 mice). (D-E) The diameter (D) and area (E) sizes of PN in WT and Rnf220+/− mice. Each data presents the average diameter and area sizes of PN from four consecutive slices for completely presenting circular fluorescence projections. (F) The area of fluorescence projection from motor cortex to PN in WT and Rnf220+/− mice. The sample used for statistics is consistent with the one selected in Figure D-E. Each data represents the average fluorescence area of four consecutive slices. (G) The proportion of projected-fluorescence area from motor cortex to PN area. The sample used for statistics is consistent with the one selected in Figure D-E. WT, wild-type. HE, heterozygote. n.s., not significant. **p < 0.01.

There is no difference of structure in PN or even pons between WT and Rnf220+/−mice (Figure 2-figure supplement 2A and B; Figure 2C-E). Considering the motor deficits of Rnf220+/− mice (Ma, et al., 2021; Ma et al., 2019), the neuronal circuit between the motor cortex and the PN neurons was traced anterogradely by a non-transneuronal tracing virus, (rAAV-hSyn-EGFP-WPRE-hGH-polyA) (Figure 2B). The results showed that fewer PN neurons were targeted by axons from the motor cortex in the Rnf220+/− mice (Figure 2C and F-G; Figure 2-figure supplement 2C). In addition, the projection from the motor cortex was centralized in the PN of Rnf220+/−mice (Figure 2C; Figure 2-figure supplement 2B).

Taken together, both the expression level/pattern of the endogenous Hox genes and the neuronal projection pattern from the motor cortex were disturbed by Rnf220 insufficiency in the mouse PN.

RNF220 targets WDR5 for K48-linked polyubiquitination and degradation

The up-regulation of Hox genes in the Rnf220+/− mice suggested a role for RNF220 in the maintenance of Hox expression pattern. Generally, the Hox expression pattern is maintained epigenetically by the PcG and trxG complexes, which regulate the silencing and activation of Hox genes, respectively (Bahrampour et al., 2019; Chopra et al., 2009; Kang et al., 2022; Papp et al., 2006). Indeed, we did observe the local epigenetic modification changes, i.e. the downregulated H3K27me3 signals and upregulated H3K4me3 signals, in some Hox cluster in hindbrain of Rnf220+/− mice by ChIP-qPCR assays (Figure 3-figure supplement 1A and B). Therefore, protein levels of the core components of PcG and TrxG complexes in the mouse hindbrain were examined and the results showed that although the protein levels of the core components of PcG we examined were comparable between WT and Rnf220+/−, a clear increase in the protein level of WDR5, a key component of the trxG complex, was observed in the hindbrain of both Rnf220+/−and Rnf220−/− mouse embryos at E18.5 (Figure 3A; Figure 3-figure supplement 2A and B), the phenomenon was also observed at E16.5 (Figure 3B). We next tested the expression of WDR5 in the mouse pons, cortex, and cerebellum at adult and found that the increase in the protein level of WDR5 was only observed in the pons, but not in the cortex or cerebellum, in Rnf220+/− mice (Figure 3C; Figure 3-figure supplement 2C and D), which is consistent with the pons-specific up-regulation in Hox genes expression. In addition, the protein level of WDR5 was also enhanced in the PN of Rnf220+/− mice (Figure 3D) and Rnf220 knockdown P19 cells in the presence of RA (Figure 3E).

RNF220 mediates WDR5 degradation.

(A-D) Western blots analysis showing the protein level of WDR5 in the indicated brain tissues of mice with different genotypes at different ages. (E) Western blots showed WDR5 levels of in the pons of adult mice with indicated genotypes. (F) Western blot analysis of protein levels of WDR5 in P19 cells with Rnf220 knockdown or not in the presence or absence of RA. IB, immunoblot. WT, wild-type. HE, heterozygote. KO, knockout. PN, pontine nuclei. NC, negative control. RA, retinoic acid.

The above data suggest that WDR5 might be a direct target of RNF220 for polyubiquitination and thus degradation. Indeed, in HEK293 cells transiently co-transfected with Flag-tagged RNF220 and myc-tagged WDR5, WDR5 coimmunoprecipitated with RNF220 (Figure 4A). In the reverse experiment, RNF220 also coimmunoprecipitated with WDR5 (Figure 4B). Furthermore, when WDR5 was immunoprecipitated from the mouse brainstem using an anti-WDR5 antibody, endogenous RNF220 was also detected in the immunoprecipitated (Figure 4C). Then we examined if the WDR5 protein level was regulated by RNF220. The results showed that co-expression of wild-type RNF220, but not the ligase-dead mutant (W539A) or the RING domain deletion (△Ring) form, clearly reduced the protein level of WDR5 in HEK293 cells (Figure 4D). In addition, the reduction in the protein level of WDR5 by RNF220 overexpression was blocked by MG132, suggesting a role for the proteasome in this regulation (Figure 4E).

RNF220 interacts with and targets WDR5 for K48-linked polyubiquitination.

(A-B) Co-immunoprecipitation (co-IP) analysis of interactions between RNF220 and WDR5 in HEK293 cells. HEK293 cells were transfected with indicated plasmids and harvested after 48 h. Cell lysates were immunoprecipitated with anti-FLAG beads. Whole-cell lysate and immunoprecipitates were subjected to western blot analysis using indicated antibodies. (C) Endogenous co-immunoprecipitation analysis showing the interaction between RNF220 and WDR5 in hindbrains of WT mice. (D) Western blots analysis shows the protein level of WDR5 when co-expressed with wild-type or mutated RNF220 in HEK293 cells. (E) Western blots analysis shows the protein level of WDR5 when co-expressed with RNF220 in HEK293 cells in the presence of MG132 (10 mM) or not. (F) In vivo ubiquitination assays showing the ubiquitination status of WDR5 when co-expressed with WT or mutated RNF220 in HEK293 cells. (G) In vivo ubiquitination assays showing the ubiquitination status of WDR5 in hindbrains of WT and Rnf220+/− mice. (H) In vivo ubiquitination assays showing RNF220-induced polyubiquitination of WDR5 when the indicated ubiquitin mutations were used in HEK293 cells. (I) In vivo ubiquitination assays showing the ubiquitination status of the indicated WDR5 mutants when co-expressed with WT or ligase dead RNF220 in HEK293 cells. WT, wild-type. HE, heterozygote. KO, knockout. IB, immunoblot. IP, immunoprecipitation. UB, ubiquitin. WCL, whole cell lysate. △Ring, RNF220 Ring domain deletion. W539R, RNF220 ligase dead mutation. K48, ubiquitin with all lysines except the K48 mutated to arginine. K48R, ubiquitin with the K48 was substituted by an arginine. 3KR, substitution of lysines at the positions of 109, 112, and 120 in WDR5 with arginines simultaneously.

We further carried out polyubiquitination assays to examine if WDR5 may be a target for RNF220. It was found that co-expression of RNF220 strongly enhanced the polyubiquitination of WDR5 protein, and this regulation depends on the E3 ubiquitin ligase activity of RNF220 because RNF220 △ Ring failed to promote the polyubiquitinaion of WDR5 (Figure 4F). Moreover, when the endogenous WDR5 proteins were immunoprecipitated from the mouse pons, its polyubiquitination level in Rnf220+/− mice was markedly decreased compared to WT mice (Figure 4G).

Different types of ubiquitination linkages have distinct regulatory effects on the stability or activity of protein targets, and K48-linked polyubiquitination usually leads to proteasomal degradation of protein targets (Grice et al., 2016). Indeed, WDR5 were only ubiquitinated by RNF220 when K48-type ubiquitin was present (Figure 4H). Consistent with this, when the K48R mutated ubiquitin was co-expressed, the RNF220 induced polyubiquitination level of WDR5 protein was fully diminished (Figure 4H). To determine the exact lysines ubiquitinated by RNF220, we first tested the effects of RNF220 on the stability of different WDR5 truncates and found that the truncate remaining 1-127aa was enough to be degraded by RNF220, suggesting the corresponding lysines were included in this region (Figure 4-figure supplement 1A). We individually mutated all these conserved lysines into arginines and then examined RNF220-mediated ubiquitination of each mutant. It was found that K109, K112, and K120 were required for the polyubiquitination of WDR5 (Figure 4-figure supplement 1B). Furthermore, when these lysine residues were simultaneously mutated into arginines, RNF220 failed to enhance the polyubiquitination levels of the resulted WDR5-3KR mutants (Figure 4I), suggesting that these lysine residues are direct ubiquitination sites. Together, these results indicate that RNF220 regulates WDR5 ubiquitination by adding K48-linked polyubiquitin chains at the lysine sites of K109, K112, and K120.

The maintenance of Hox expression by RNF220 depends on its regulation to WDR5

To investigate the requirement of the RNF220-mediated regulation to WDR5 in Hox expression maintenance by RNF220, we first examined whether Wdr5 knockdown could mitigate the impact of Rnf220 knockdown on Hox expression in P19 cells in the presence of RA. It is observed that the stimulation of Hox genes by Rnf220 knockdown was significantly reduced when Wdr5 was knocked-down simultaneously by siRNAs in the presence of RA (Figure 5A and B). Note that the expression levels of Hox genes showed no clear changes when without RA or only Wdr5 was knocked-down in P19 cells (Figure 5-figure supplement 1A and B).

WDR5 recovered Rnf220 deficiency-induced up-regulation of Hox genes in P19 cell line.

(A-B) qRT-PCR analysis showing the expression levels of Rnf220 (A) and Wdr5 (B) when transfected the indicated combinations of siRNAs against Rnf220 or Wdr5 in the presence or absence of RA. Bar graphs show the relative levels normalized against control group without siRNA or RA treatment. (C) qRT-PCR analysis showing the expression levels of HoxA1, HoxB1, HoxA9, HoxB9 when transfected siRnf220 or both siRnf220 and siWdr5 when treated with RA. RA: retinoic acid. n.s., not significant. ***p < 0.001.

Using in-utero microinjection, we tested the effects of WDR5 inhibition by WDR5-IN-4, an established WDR5 inhibitor (Aho et al., 2019), on the upregulation of Hox genes in hindbrains of the Rnf220+/− mice at late embryonic stages (Figure 6A). It was found that the up-regulation of Hox3-5 genes in the hindbrain of the Rnf220+/−embryos was largely reversed by WDR5-IN-4 (Figure 6C). Notably, we found that WDR5-IN-4 injection had no effect on the endogenous Rnf220 expression in the hindbrain (Figure 6B). Last, the effect of genetic ablation of Wdr5 in neural system on the upregulation of Hox genes in RNF220 hypoinsufficient mice was examined. Considering Wdr5+/− mice is embryonic lethal, and the up-regulation of Hox genes could be also observed in the hindbrains of Rnf220fl/wt;Nestin-Cre mice, we used Rnf220 fl/wt;Nestin-Cre and Wdr5fl/wt mice to generate WT, Rnf220 knockdown; Wdr5 knockdown and Rnf220 andWdr5 double knockdown mice (Figure 6D). It was found that the upregulation of Hox3-5 genes in the pons of Rnf220 hypoinsufficient mice were markedly recovered by Wdr5 genetic ablation (Figure 6E). ChIP-qPCR also showed the local epigenetic modification changes in some Hox cluster, with H3K27me3 upregulated and H3K4me3 downregulated in Rnf220 and Wdr5 double knockdown P19 cell line related to Rnf220 single knock down P19 cell line in the presence of RA (Figure 6-figure supplement 1).

Genetic and pharmacological ablation of WDR5 recovered Rnf220 deficiency-induced up-regulation of Hox genes.

(A) Diagram of experimental strategy for in utero local injection of WDR5 inhibitors. (B-C) qRT-PCR analysis of expression levels of Rnf220 (B), Hox3-Hox5 (C) in hindbrains of WT and Rnf220+/−mouse embryos treated with WDR5 inhibitors or not at E18.5 (n=3 mice per group). Actin was used as the internal controls. Heatmap of Hox expression showed the relative levels normalized against WT group. (D-E) qRT-PCR analysis of expression levels of Rnf220 (D), Wdr5 (D), Hox3-Hox5 (E) in pons of P15 mice with indicated genotypes (n=2 mice per group). GAPDH was used as the internal controls. Heatmap of Hox expression showed the relative levels normalized against WT group. WT, wild-type. HE, heterozygote.

Taken together, the above findings support the involvement of WDR5 regulation in RNF220 mediated the maintenance of Hox expression pattern in the mouse hindbrain.

Discussion

During early embryonic development, the Hox genes are collinearly expressed in the hindbrain and spinal cord along the A-P axis to guide regional neuronal identity. Later on, the segmental Hox gene expression pattern in the hindbrain is maintained till at least early postnatal stages. The Hox genes are also expressed in adult hindbrains with restricted anterior boundaries. (Philippidou et al., 2013; Miller et al., 2024; Bonito et al., 2013). In addition to their roles in progenitor cell specification, cell survival, neuronal migration, axon guidance, and dendrite morphogenesis during early development stages (Smith et al., 2024), Hox genes also play key roles in the regulation of synapse formation, neuronal terminal identity and neural circuit assembly at late stages (Feng et al., 2020; Feng et al., 2021; Philippidou et al., 2013). Our study revealed a dose-dependent role of RNF220 and WDR5 in the maintenance of Hox expression in the hindbrain, which might have a functional role in the neural circuit organization of the pons in mice.

In mammals, WDR5, a key component of the COMPASS-related complexes, has been reported to be absolutely required for the regulation of Hox genes expression during mammalian embryo development and cancer progression via different mechanisms (Wysocka et al., 2005;Chen et al., 2021; Yu et al., 2021). Here, we report a WDR5 protein stability controlling mechanism by RNF220-mediated polyubiquitination and illustrate the role of this regulation on the maintenance of Hox genes expression in the mouse hindbrain. The stabilization of WDR5 and stimulation of Hox expression upon RNF220 knockdown was also recapitulated in RA treated P19 cells, a cellular model for Hox regulation study.

Interestingly, when examined at single cell level, different Hox stimulation profiles were detected in different neuronal groups, so that Hox 1-3 were up-regulated in cluster 10, while Hox3-5 and Hox7-10 upregulation appeared in cluster 7 and cluster 2 respectively (Figure 1D). Cluster 10 and 7 express endogenous Hox1-3 and Hox3-5, respectively. It is reasonable that the expression of these Hox genes are stimulated upon WDR5 elevation due to RNF220 insufficiency. Cluster 2 is most interesting in that, Hox7-10 were up-regulated in these cells which were not expressed in wild type hindbrain. This cluster was defined as glycinergic neuron cells with Slc6a5 expression. The nature of this group of cells and the phenotypic effect of Hox up-regulation await further analysis. The de-regulation of Hox genes might have wide effect on the development of the many nuclei in the pons, including cell differentiation and neural circuit formation. For the PN, we showed here that the patterning and projection pattern from the motor cortex were affected in the Rnf220+/− mice.

The regulation of WDR5 and Hox expression by RNF220 seems to be context dependent and precisely controlled in vivo, depending on the molecular and epigenetic status of the cell. For example, although WDR5 and RNF220 are widely co-expressed in the brain, WDR5 levels are not affected in most regions in the Rnf220+/− brain, including the cortex, cerebellum etc. (Figure 3-figure supplement 2C and D). This is also true for EED, another epigenetic regulator, which is targeted by RNF220 in the cerebellum, but not in the pons (Ma et al.,2020). In untreated P19 cells, knockdown of RNF220 also has no effect on the stability of endogenous WDR5, although overexpression of RNF220 did reduce WDR5 level in 293T cells. We suggest that RA treatment of P19 cells helps to set up the molecular environment that facilitates Hox expression as well as RNF220-WDR5 interaction.

In summary, our data support WDR5 as a RNF220 target involved in the maintenance of Hox expression and thus development of the pons.

Materials and methods

Mouse strains and genotyping

All procedures involving mice were conducted in accordance with the guidelines of the Animal Care and Use Committee of the Kunming Institute of Zoology, Chinese Academy of Sciences (IACUC-PA-2021-07-018). Mice were housed under standard conditions at a temperature range of 20–22 °C, humidity of 30%–45%, and 12-h light/dark cycle. Rnf220 floxed mice (Rnf220fl/fl) were originally 129Sz/SvPasCrl background and mated with C57BL/6 then. Other mice used were maintained on a C57BL/6 background. Vasa-Cre mice were used to mate with Rnf220 floxed mice (Rnf220fl/fl) to generate Rnf220 germ cell knockout (Rnf220−/−) or heterozygote (Rnf220+/−) mice. Nestin-Cre mice were used to mate with Rnf220fl/fl and Wdr5 floxed mice (Wdr5fl/fl) to generate conditional neural specific knockout or heterozygote mice. Rnf220fl/wt;Wdr5fl/wt;Nestin-Cre mice were obtained by crossing Rnf220fl/wt;Nestin-Cre with Wdr5fl/wt mice.

Genotypes were determined by PCR using genome DNA from tail tips as templates. PCR primers were listed as follows: 5’-CTG TTG ATG AAG GTC CTG GTT-3’ and 5’-CAG GAA AAT CAA TAG ACA ACT T-3’ were used to detect Rnf220 floxP carrying. 5’-CTG TTG ATG AAG GTC CTG GTT-3’ and 5’-CTG ATT TCC AGC AAC CTA AA-3’ were used to detect Rnf220 knockout. 5’-GCC TGC ATT ACC GGT CGA TGC-3’ and 5’-CAG GGT GTT ATA AGC AAT CCC-3’ were used for Cre positive detection. 5’-GAA TAA CTA CTT TCC CTC AGA CC-3’ and 5’-CAG GCC AAG TAA CAG GAG GTA G-3’ were used to detect Wdr5 floxP carrying. 5’-GAA TAA CTA CTT TCC CTC AGA CC-3’ and 5’-AGA CCC TGA GTG AGG ATA CAT AA-3’ were used to detect Wdr5 knockout.

Cell culture

HEK293 and P19 cell lines were grown in Dulbecco’s Modified Eagle Medium (DMEM) (Gibco, C11995500BT) supplementary with 10% fetal bovine serum (FBS) (Biological Industries, 04-001-1A), 100 units/mL penicillin and 100 mg/ml streptomycin (Biological Industries, 03-031-1B). Cell cultures were maintained at 37°C in a humidified incubator with 5% CO2. 0.5μM RA (Sigma, R2625) was used to induce Hox expression in P19 cells.

To achieve gene overexpression or knockdown, HEK293 and P19 cells were transfected by Lipo2000 (Invitrogen, 11668500) according to the manufacturer’s instructions. The following small interfering RNAs (siRNAs) (RiboBio) were used for Rnf220 or Wdr5 knockdown in P19 cells: siG2010160325456075, siG2010160325457167 and siG2010160325458259 were used for Rnf220 knockdown; siB09924171210, siG131113135429 and siG131113135419 were used for Wdr5 knockdown.

Total RNA isolation and qRT–PCR

Tissue and cells were homogenized with 1 mL TRIZOL (TIANGEN, DP424), after which 200 μL of chloroform was added to the lysates for phase separation. After centrifugation at 12L000 g for 15 min at 4 °C, the aqueous phase (500 μL) was transferred to a new tube and mixed with equal volumes of isopropanol for RNA precipitation. After 30 min, RNA pellets were harvested by centrifugation at 12L000 g for 15 min at 4 °C, twice washed with 75% ethanol, and dissolved in DNase/RNase-free water.

cDNA was synthesized with Strand cDNA Synthesis Kit (ThermoScientific, K1632) according to the manufacturer’s instructions. All reactions were performed at least triplicates with Light Cycler 480 SYBR Green L Master (Roche, 04707516001). Primers used for real-time polymerase chain reaction list in Table 4.

ChIP-qPCR

P19 cells or grinded mouse PN tissues were cross-linked with 1% formaldehyde for 25 min, and stopped with addition of glycine. After gentle centrifuging, samples were collected and lysed with ChIP lysis buffer (150 mM NaCl, 25 mM Tris, pH 7.5, 1% Triton X-100, 0.1% SDS, 0.5% deoxycholate, and complete protease inhibitor cocktail) for 30 min on ice. Genomic DNA was sheared into 200-500 bp fragments by sonication. After centrifugation, the supernatants containing chromatin fragments were collected and immunoprecipitated with anti-IgG (2 μg, ProteinTech, B900610), anti-H3K27me3 (2 μg, Abcam, ab192985), or anti-H3K4me3 antibodies (2 μg, Abcam, ab8580) at 4L overnight. Then the immunoprecipitates coupled with protein A/G agarose beads (Santa Cruz, sc-2003) were washed sequentially by a low salt washing buffer, high salt washing buffer, LiCl washing buffer, and TE buffer. The immunoprecipitated chromatin fragments were eluted by 500 μL elution buffer for reversal of cross-linking at 65 °C overnight. Input or immunoprecipitated genomic DNA was purified by the QIAquick PCR Purification Kit (Qiagen, 28104) and used as a template for quantification PCR. The primers we used were listed in Table 5.

In utero microinjection

Pregnant mice were administered isoflurane for deep anesthesia. Following this, a laparotomy was performed to carefully extract the embryos, which were then placed on a sterile surgical drape. WDR5-IN-4 (100 μg, MedChemExpress, HY-111753A), containing 0.05% Malachite Green reagent for tracing, was injected into the hindbrain of E15.5 embryos using a finely-tapered borosilicate needle. To minimize the risk of spontaneous abortion, injections were spaced for the selected embryos. Following the injections, the uterus was returned to the abdominal cavity and infused with 2 mL of 37 °C, 0.9% saline solution. The peritoneum and abdominal skin were then sutured. Finally, the mice were placed on a heating pad to facilitate recovery from anesthesia.

Ubiquitination assay, immunoprecipitation and western blot analysis

In vivo ubiquitination, immunoprecipitation and Western blot assays were carried out as previously described (Ma et al., 2023.). HEK293 cells were transfected in 6-well plates, and at 48 h post transfection cells were lysed in 500 μl of lysis buffer (50 mM Tris-HCl [pH 7.4], 150 mM NaCl, 5 mM EDTA [pH 8.0], and 1% Triton X-100) that contained a protease inhibitor mixture (Roche Applied Science) for 30 min on ice; mouse tissues were fully grinded and lysed with RIPA Strong Lysis (Beyotime, P0013B) containing a protease inhibitor mixture for 30 min on ice. Following centrifugation at 12,000 rpm for 10 min at 4°C, 50μl supernatant was mixed with loading buffer and incubate boiled at 95°C for 5 min, the remaining supernatant was coating with antibody and A/G agarose beads (or FLAG beads) overnight. After washing 5 times, the bound proteins were eluted with SDS loading buffer at 95°C for 5 min. Final total lysates and immunoprecipitates were subjected to SDS-PAGE and Western blot analysis. The following primary antibodies were used for botting: anti-RNF220 (Sigma-Aldrich, HPA027578), anti-WDR5 (D9E11) (Cell Signaling Technology, 13105), anti-Ring1B (D22F2) (Cell Signaling Technology, 5694S), anti-Sin3B (AK-12) (Santa Cruz, sc-768), anti-EZH2 (ProteinTech, 21800-1-AP), anti-SUZ12 (Bethyl, A302-407A), anti-RYBP (A-1) (Santa Cruz, sc-374235), anti-CBX6 (H-1) (Santa Cruz, sc-393040), anti-CBX7 (G-3) (Santa Cruz; sc-376274), anti-CBX8 (C-3) (Santa Cruz, sc-374332), anti-PHC1 (D-10) (Santa Cruz, sc-390880), anti-α-Tubulin (ProteinTech, 66031-1-Ig), anti-FLAG (Sigma-Aldrich; F-7425), anti-myc (ProteinTech; 16286-1-AP), and anti-Ub (P4D1) (Santa Cruz, sc-2007)

snRNA-seq library preparation, sequencing, and data analysis

snRNA-seq libraries were prepared using the Split-seq platform (Butler et al., 2018; Rosenberg et al., 2018). Freshly harvested mouse pons tissues underwent nuclear extraction following previously described protocols (Butler et al., 2018; Rosenberg et al., 2018.). In brief, mouse brain tissues were transferred into a 2 mL Dounce homogenizer containing 1mL homogenizing buffer (250 mM sucrose, 25 mM KCl, 5 mM MgCl2, and 10 mM Tris-HCl [pH = 8.0], 1 mM DTT, RNase Inhibitor and 0.1% Triton X-100). The mixture was subjected to five strokes with a loose pestle, followed by 10 strokes with a tight pestle. The resulting homogenates were filtered through a 40-μm strainer into 5-mL Eppendorf tubes and subsequently centrifuged for 4 min at 600 g and 4 °C. The pellet was re-suspended and washed in 1 mL of PBS containing RNase inhibitors and 0.1% BSA. The nuclei were again filtered through a 40-μm strainer and quantified. These nuclei were partitioned into 48 wells, each of which contained a barcoded, well-specified reverse transcription primer, to enable in-nuclear reverse transcription. Subsequent second and third barcoding steps were carried out through ligation reactions. After the third round of barcoding, the nuclei were divided into 16 aliquots and lysed prior to cDNA purification. The resulting purified cDNA was subjected to template switching and qRT-PCR amplification, which was halted at the beginning of the plateau stage. Finally, the purified PCR products (600 pg) were used to generate Illumina-compatible sequencing libraries. Distinct, indexed PCR primer pairs were employed to label each library, serving as the fourth barcode.

The libraries underwent sequencing on the NextSeq system (Illumina) using 150-nucleotide kits and paired-end sequencing protocols. In this arrangement, Read 1 covered the transcript sequences and Read 2 contained the UMI and UBC barcode combinations. Initially, a six-nucleotide sequencing index, serving as the fourth barcode, was appended to the ends of Read 2. Subsequently, reads with more than one mismatching base in the third barcode were excluded from the dataset. Furthermore, reads containing more than one low-quality base (Phred score ≤ 10) within the UMI region were also discarded. The sequencing results were aligned to exons and introns in the reference genome (Genome assembly GRCm38, https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000001635.20/) and aggregated intron and exon counts at the gene level were calculated by kallisto and bustools software as described (https://bustools.github.io/BUS_notebooks_R).

Following export of the matrix, quality control measures were performed to remove low-quality cells and potential doublets, as described previously (Rosenberg et al., 2018.). After filtering, a total of 125,956 and 122,266 cells for wild type and Rnf220+/−pons respectively remained for subsequent analysis. Seurat v2 (Butler et al., 2018.) was used to identify HVGs within each group. Principal component analysis (PCA) and UMAP were performed to embed each cell from the reduced principal component space on a 2D map. Then clustering of cell populations and identification of differentially expressed genes were carried out as previously described (Ma et al., 2022.). We annotated the embryonic cell populations and lineages based on their DEGs (Table 3).

Mouse brain stereotactic injection and neuronal tracing

Mouse brain stereotactic injection and neuronal tracing were carried out as previously described with minor modifications (Xu et al., 2021.). Adult mice (2–3 months old) were used for anterograde monosynaptic neuronal tracing. The mice were first deeply anesthetized with isoflurane and placed into a stereotactic apparatus with the front teeth latched onto the anterior clamp. The mouth and nose of the mice were covered with a mask to provide isoflurane and keep them in an anesthetic state during the operation. The head was adjusted and maintained in horizontal by inserting ear bars into the ear canal. The hair of the head was shaved with an electric razor and cleaned using a cotton swab which was dipped in 75% alcohol. Cut the scalp along the midline with surgical scissors and make sure the bregma, lambda and the interaural line were exposed. The intersection between the lambda and interaural line was set to zero. The coordinates ±1.75, −4.9, −1.65) were applied for motor cortex localization. Each mouse received a single injection of 500 nL of rAAV-hSyn-EGFP-WPRE-hGH-poly A (Brainvta, PT-1990) viral fluid in the hemisphere. After the injection, mice were bred for another 3 weeks for neuronal tracing before their brains were harvested for analysis.

Slice preparation

Mice subjected to stereotactic injection were euthanized, and whole brains were collected for the preparation of frozen sections. Following fixation in 4% paraformaldehyde (diluted in PBS) for 48 h, the brains were successively treated with 20% sucrose (diluted in PBS) for 48 h and 30% sucrose for 24 h. Brain tissues were then embedded in Optimal Cutting Temperature (OTC) compound (SAKURA, 4583) at −20 L and sliced along the sagittal plane at a thickness of 40 μm.

Quantification and statistical analysis

PN and fluorescence length and area were statistically analyzed by Image J software (National Institute of Health). Statistical analyses were performed using GraphPad Prism software (GraphPad Software Inc., La Jolla, CA, USA). Comparisons were performed using the two-tailed Student’s t-test. p-values of less than 0.05 were considered statistically significant (*), 0.01 were considered statistically significant (**) and 0.001 were considered statistically significant (***).